---
title: "Figure for Institut Curie"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
    html_document:
      theme: cerulean
      toc: true
      toc_float: true
      number_sections: true
      self_contained: true
---

```{r}
library(tidyverse)
library(readxl)
library(VennDiagram)
library(grid)
source('sgRNA_library_refining_functions.R')
```

```{r}
brunello <- 'broadgpp-brunello-library-contents_full_report.xlsx'
toronto <- 'tkov3_guide_sequence_full_report.xlsx'
cellecta_M1 <- 'sample_M1_150K_full_report.xlsx'
cellecta_M2 <- 'sample_M2_150K_full_report.xlsx'
cellecta_M3 <- 'sample_M3_150K_full_report.xlsx'

brunello_report <- read_excel(paste('./output_cleaning/', brunello, sep = ''))
toronto_report <- read_excel(paste('./output_cleaning/', toronto, sep = ''))
cellecta_report_M1 <- read_excel(paste('./output_cleaning/', cellecta_M1, sep = ''))
cellecta_report_M2 <- read_excel(paste('./output_cleaning/', cellecta_M2, sep = ''))
cellecta_report_M3 <- read_excel(paste('./output_cleaning/', cellecta_M3, sep = ''))

cellecta_report <- rbind(cellecta_report_M1, cellecta_report_M2, cellecta_report_M3)
cellecta_report <- cellecta_report %>%
    arrange() %>%
    distinct()

```

```{r}

calculate_percentage <- function(library_report, discarded_sgrna, original_sgrna) {
    percentage_discarded <- (sum(library_report[[discarded_sgrna]])/sum(library_report[[original_sgrna]])) * 100
    return(percentage_discarded)
}

cellecta_multi_target <- round(calculate_percentage(cellecta_report, "multi-target sgRNA", "sgRNA number"), 2)
brunello_multi_target <- round(calculate_percentage(brunello_report, "multi-target sgRNA", "sgRNA number"), 2)
toronto_multi_target <- round(calculate_percentage(toronto_report, "multi-target sgRNA", "sgRNA number"), 2)

cellecta_single_mismatch <- round(calculate_percentage(cellecta_report, "single mismatch sgRNA", "sgRNA number"), 2)
brunello_single_mismatch <- round(calculate_percentage(brunello_report, "single mismatch sgRNA", "sgRNA number"), 2)
toronto_single_mismatch <- round(calculate_percentage(toronto_report, "single mismatch sgRNA", "sgRNA number"), 2)

cellecta_pam_distal <- round(calculate_percentage(cellecta_report, "PAM distal double mismatch", "sgRNA number"), 2)
brunello_pam_distal <- round(calculate_percentage(brunello_report, "PAM distal double mismatch", "sgRNA number"), 2)
toronto_pam_distal <- round(calculate_percentage(toronto_report, "PAM distal double mismatch", "sgRNA number"), 2)

```

# create dataframe for multi-target, single-mismatch, and pam-distal from all library

```{r}
multi_target <- c(brunello_multi_target, toronto_multi_target, cellecta_multi_target)
single_mismatch <- c(brunello_single_mismatch, toronto_single_mismatch, cellecta_single_mismatch)
pam_distal <- c(brunello_pam_distal, toronto_pam_distal, cellecta_pam_distal)

library <- c("brunello", "toronto_v3", "cellecta")
factor <- c(2,3,1)

sum_dataframe <- data.frame(library, multi_target, single_mismatch, pam_distal)

```

# Histogram of multi-target/off-target sgRNAs in Cellecta, Brunello, and Toronto

```{r}

# Reorder the library factor to ensure the bars are ordered as Cellecta, Brunello, and Toronto
sum_dataframe_long <- sum_dataframe %>%
    pivot_longer(cols = c("multi_target", "single_mismatch", "pam_distal"), 
                 names_to = "category", 
                 values_to = "percentage") %>%
    mutate(category = factor(category, levels = c("multi_target", "single_mismatch", "pam_distal")),
           library = factor(library, levels = c("cellecta", "brunello", "toronto_v3")))

# Define custom colors for the libraries
custom_colors <- c("cellecta" = "lightgreen", "brunello" = "powderblue", "toronto_v3" = "mistyrose")

# Create the grouped multibar chart
multi_bar_chart <- ggplot(sum_dataframe_long, aes(x = category, y = percentage, fill = library)) +
    geom_bar(stat = "identity", position = position_dodge2(reverse = TRUE, padding = 0.2), width = 0.7) +
    theme_bw() +
    labs(title = "", x = "", y = "Percentage of sgRNAs (%)") +
    scale_x_discrete(labels = c('multi_target' = 'Multi-target sgRNA',
                                'single_mismatch' = 'Single mismatch sgRNA',
                                'pam_distal' = 'PAM distal \n double mismatch sgRNA')) +
    geom_text(aes(label = percentage), position = position_dodge2(width = 0.7, reverse = TRUE), vjust = -0.5, size = 12) +
    theme(axis.text.x = element_text(size = 40, vjust = 0.7, colour = 'black'),
          axis.text.y = element_text(size = 40),
          axis.title.y = element_text(size = 40, vjust = 0.5, margin = margin(r = 20)),
          legend.title = element_text(size = 40),
          legend.text = element_text(size = 40),
          plot.margin = margin(1, 1, 1, 1), 
          line = element_line((size = 4), colour = 'black'),
          legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = custom_colors,
                      name = 'Library',
                      breaks = c("cellecta", "brunello", "toronto_v3"),
                      labels = c("cellecta" = "Cellecta", 
                                 "brunello" = "Brunello", 
                                 "toronto_v3" = "TKoV3"))

multi_bar_chart

ggsave(
    path = './',
    width = 30,
    height = 15,
    dpi = 300,
    plot = multi_bar_chart,
    filename = paste('multi_bar_chart_off_target.png', sep = ''))
```


# Number of genes completely eliminated due to the multi-target sgRNAs

```{r}
hgnc <- read_tsv(url('https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt'))

hgnc_selected <- hgnc %>%
  select(symbol, name, locus_group, locus_type) %>%
  dplyr::rename('gene' = 'symbol')

genes_completely_lost <- function(library_report, sgrna_discarded, original_sgrna) {
    data <- library_report %>%
        select(gene, original_sgrna, sgrna_discarded) %>%
        filter(.data[[original_sgrna]] == .data[[sgrna_discarded]]) %>%
        inner_join(hgnc_selected, join_by(gene))
    
    return(data)
}

genes_lost_multi_target_cellecta <- genes_completely_lost(cellecta_report, 'multi-target sgRNA', 'sgRNA number')
genes_lost_multi_target_brunello <- genes_completely_lost(brunello_report, 'multi-target sgRNA', 'sgRNA number')
genes_lost_multi_target_toronto <- genes_completely_lost(toronto_report, 'multi-target sgRNA', 'sgRNA number')

## ---------------------------------------------------------------

genes_lost_single_mismatch_cellecta <- genes_completely_lost(cellecta_report, 'single mismatch sgRNA', 'sgRNA number')
genes_lost_single_mismatch_brunello <- genes_completely_lost(brunello_report, 'single mismatch sgRNA', 'sgRNA number')
genes_lost_single_mismatch_toronto <- genes_completely_lost(toronto_report, 'single mismatch sgRNA', 'sgRNA number')

## ---------------------------------------------------------------

genes_lost_pam_distal_cellecta <- genes_completely_lost(cellecta_report, 'PAM distal double mismatch', 'sgRNA number')
genes_lost_pam_distal_brunello <- genes_completely_lost(brunello_report, 'PAM distal double mismatch', 'sgRNA number')
genes_lost_pam_distal_toronto <- genes_completely_lost(toronto_report, 'PAM distal double mismatch', 'sgRNA number')

# ---------------------------------------------------------------------

genes_lost_multi_target <- c(length(genes_lost_multi_target_brunello$gene), length(genes_lost_multi_target_toronto$gene), length(genes_lost_multi_target_cellecta$gene))
genes_lost_single_mismatch <- c(length(genes_lost_single_mismatch_brunello$gene), length(genes_lost_single_mismatch_toronto$gene), length(genes_lost_single_mismatch_cellecta$gene))

sum_dataframe_genes <- data.frame(library, genes_lost_multi_target, genes_lost_single_mismatch, factor)

```

# barplot for genes lost

```{r}
barplot <- ggplot(sum_dataframe_genes, aes(x = reorder(library, factor), y = genes_lost_multi_target)) +
    geom_histogram(stat = "identity", fill = c("powderblue", "mistyrose", "lightgreen")) +
    theme_minimal() +
    labs(title = "", x = "", y = "Number of genes") +
    scale_x_discrete(labels=c('brunello' = 'Brunello \n 77K sgRNAs',
                              'cellecta' = 'Cellecta \n 150K sgRNAs',
                              'toronto v3' = 'Toronto \nKnockout V3 \n 71K sgRNAs')) +
    geom_text(aes(label = genes_lost_multi_target), vjust=-0.5, size = 10) +
    theme(axis.text.x = element_text(size = 30),
        axis.text.y = element_text(size = 30),
        axis.title.y = element_text(size = 30, vjust=0.5, margin = margin(r = 20)))
barplot

```

# venn diagram for genes lost

```{r}
single_mismatch_venn <- venn.diagram(
    x = list(
        genes_lost_single_mismatch_brunello$gene,
        genes_lost_single_mismatch_toronto$gene,
        genes_lost_single_mismatch_cellecta$gene
    ),
    category.names = c("Brunello \n 77K sgRNAs", "Toronto Knockout V3 \n 71K sgRNAs", "Cellecta \n 150K sgRNAs"),
    filename = NULL,
    output = TRUE,
    fill = c("powderblue", "mistyrose", "lightgreen"),
    alpha = 0.5,
    cex = 2,
    cat.cex = 2,
    cat.pos = c(-20, 20, 180),
    cat.dist = c(0.1, 0.1, 0.1)
)

ggsave(
    path = './',
    width = 12,
    height = 10,
    dpi = 300,
    plot = single_mismatch_venn,
    filename = paste('VennDiagram_single_mismatch.png', sep = ''))

```

# looking at the alignment data

```{r}

multi_target_commonality <- intersect(intersect(genes_lost_multi_target_brunello$gene, genes_lost_multi_target_toronto$gene), genes_lost_multi_target_cellecta$gene)

brunello_alignment <- read_csv('./object_intermediate/broadgpp-brunello-library-contents_aln.csv')
toronto_alignment <- read_csv('./object_intermediate/tkov3_guide_sequence_aln.csv')
cellecta_M1_alignment <- read_csv('./object_intermediate/sample_M1_150K_aln.csv')
cellecta_M2_alignment <- read_csv('./object_intermediate/sample_M2_150K_aln.csv')
cellecta_M3_alignment <- read_csv('./object_intermediate/sample_M3_150K_aln.csv')

brunello_multi_target_commonality <- brunello_alignment %>%
    filter(gene %in% multi_target_commonality)

toronto_multi_target_commonality <- toronto_alignment %>%
    filter(gene %in% multi_target_commonality)

```

# Number of genes eliminated after refining

```{r}

calculate_percentage_genes_lost <- function(library_report) {
    original_gene <- length(unique(library_report$gene))
    
    genes_lost <- library_report %>%
        select(gene, `quality check`) %>%
        filter(`quality check` == 'fail')
    
    discarded_gene <- length(unique(genes_lost$gene))

    percentage_discarded <- (discarded_gene/original_gene) * 100
    return(list(percentage_discarded, genes_lost))
}

unlist(calculate_percentage_genes_lost(cellecta_report)[1])

genes_lost_cellecta <- round(unlist(calculate_percentage_genes_lost(cellecta_report)[1]), 2)
genes_lost_brunello <- round(unlist(calculate_percentage_genes_lost(brunello_report)[1]), 2)
genes_lost_toronto <- round(unlist(calculate_percentage_genes_lost(toronto_report)[1]), 2)

genes_lost <- c(genes_lost_brunello, genes_lost_toronto, genes_lost_cellecta)

sum_dataframe_genes_lost <- data.frame(library, genes_lost)

```

```{r}
barplot <- ggplot(sum_dataframe_genes_lost, aes(x = reorder(library, factor), y = genes_lost)) +
    geom_histogram(stat = "identity", fill = c("powderblue", "mistyrose", "lightgreen")) +
    theme_minimal() +
    labs(title = "", x = "", y = "Percentage of genes") +
    scale_x_discrete(labels=c('brunello' = 'Brunello \n 77K sgRNAs',
                              'cellecta' = 'Cellecta \n 150K sgRNAs',
                              'toronto v3' = 'Toronto \nKnockout V3 \n 71K sgRNAs')) +
    geom_text(aes(label = genes_lost), vjust=-0.5, size = 10) +
    theme(axis.text.x = element_text(size = 30),
        axis.text.y = element_text(size = 30),
        axis.title.y = element_text(size = 30, vjust=0.5, margin = margin(r = 20)))
barplot
```

```{r}

unlist_genes_lost_brunello <- as.data.frame(calculate_percentage_genes_lost(brunello_report)[2])
unlist_genes_lost_toronto <- as.data.frame(calculate_percentage_genes_lost(toronto_report)[2])
unlist_genes_lost_cellecta <- as.data.frame(calculate_percentage_genes_lost(cellecta_report)[2])

genes_lost_venn <- venn.diagram(
    x = list(
        unlist_genes_lost_brunello$gene,
        unlist_genes_lost_toronto$gene,
        unlist_genes_lost_cellecta$gene
    ),
    category.names = c("Brunello \n 77K sgRNAs", "Toronto Knockout V3 \n 71K sgRNAs", "Cellecta \n 150K sgRNAs"),
    filename = NULL,
    output = TRUE,
    alpha = 0.5,
    print.mode = c('raw'),
    cex = 2,
    cat.cex = 2,
    cat.pos = c(-20, 20, 180),
    cat.dist = c(0.1, 0.1, 0.1)
)

grid.newpage()
grid.draw(genes_lost_venn)
grid.polygon(
  x = unit(c(0.5, 0.6, 0.5, 0.4), "npc"),  # Adjust coordinates for the intersection
  y = unit(c(0.5, 0.6, 0.7, 0.6), "npc"),
  gp = gpar(fill = "gold", alpha = 0.7)  # Custom color for the triple intersection
)

genes_lost_venn


ggsave(
    path = './',
    width = 12,
    height = 10,
    dpi = 300,
    plot = genes_lost_venn,
    filename = paste('test.png', sep = ''))


```

# What are the genes that completely eliminated in all library

```{r}
intersection_genes_lost <- intersect(unlist_genes_lost_brunello$gene, intersect(unlist_genes_lost_toronto$gene, unlist_genes_lost_cellecta$gene))

genes_lost_hgnc <- hgnc_selected %>%
    filter(gene %in% intersection_genes_lost) %>%
    count(locus_type) %>%
    dplyr::rename('count' = 'n') %>%
    ggplot(aes(x = locus_type, y = count)) +
        geom_histogram(stat = "identity", fill = "#bbc5c1") +
        theme_minimal() +
        labs(title = "", x = "", y = "Number of genes") +
        scale_x_discrete(labels=c('RNA, long non-coding' = 'lncRNA',
                                'complex locus constituent' = 'Complex \nconstituent loci',
                                'gene with protein product' = 'Protein-coding \ngenes',
                                'pseudogene' = 'Pseudogene',
                                'unknown' = 'Unknown')) +
        geom_text(aes(label = count), vjust=-0.5, size = 10) +
        theme(axis.text.x = element_text(size = 20),
            axis.text.y = element_text(size = 30),
            axis.title.y = element_text(size = 30, vjust=0.5, margin = margin(r = 20)))


genes_lost_hgnc

ggsave(
    path = './',
    width = 17,
    height = 15,
    dpi = 300,
    plot = genes_lost_hgnc,
    filename = paste('genes_lost_hgnc.png', sep = ''))


```

# example of gene symbol correction

```{r}

library <- c('brunello', 'brunello', 'toronto', 'toronto', 'cellecta', 'cellecta')
number_of_genes <- c(16967, 1056, 16313, 1125, 17042, 1376)
annotation <- c('match', 'mismatch', 'match', 'mismatch', 'match', 'mismatch')

library_correction_report <- data.frame(library, annotation, number_of_genes)

custom_colors <- c('match' = 'lightgreen', 'mismatch' = "mistyrose")
library_correction_report <- library_correction_report %>%
    mutate(annotation = factor(annotation, levels = c("mismatch", "match")))
    barplot <- ggplot(library_correction_report, aes(x = reorder(library, -number_of_genes), y = number_of_genes, fill = annotation)) +
        geom_bar(stat = "identity", width = 0.7) +
        theme_bw(base_line_size = 15) +
        labs(title = "", x = "", y = "Number of genes") +
        scale_x_discrete(labels=c('brunello' = 'Brunello',
                                'cellecta' = 'Cellecta',
                                'toronto' = 'TKoV3')) +
        geom_text(aes(label = number_of_genes), position = position_stack(vjust = 0.5), size = 15) +
        theme(axis.text.x = element_text(size = 50, colour = 'black'),
            axis.text.y = element_text(size = 50),
            axis.title.y = element_text(size = 50, vjust=0.5, margin = margin(r = 20)),
            legend.title = element_text(size = 50),
            legend.text = element_text(size = 50),
            line = element_line((size = 2), colour = 'black'),
            plot.margin = margin(1, 1, 1, 1)) +
        scale_fill_manual(values = custom_colors,
                        name = 'Gene symbol in sgRNA ID',
                        labels = c('match' = 'Approved symbol', 
                                    'mismatch' = 'Previous/alias symbol')) +
        theme(legend.key.size = unit(1.5, "cm")) +
        scale_y_continuous(limits = c(0, 20000), breaks = seq(0, 20000, by = 5000))

barplot

ggsave(
    path = './',
    width = 25,
    height = 15,
    dpi = 300,
    plot = barplot,
    filename = paste('histogram_exon_annotation_before.png', sep = ''))


```

# final result of gene symbol correction

```{r}
library <- c('brunello', 'toronto', 'cellecta')
number_of_genes <- c(17927, 17419, 17959)
annotation <- c('match', 'match', 'match')

library_correction_report <- data.frame(library, annotation, number_of_genes)

custom_colors <- c('match' = 'lightgreen', 'mismatch' = "mistyrose")

barplot <- ggplot(library_correction_report, aes(x = reorder(library, -number_of_genes), y = number_of_genes, fill = annotation)) +
    geom_bar(stat = "identity", width = 0.7) +
    theme_bw() +
    labs(title = "", x = "", y = "number of genes") +
    scale_x_discrete(labels=c('brunello' = 'Brunello',
                              'cellecta' = 'Cellecta',
                              'toronto' = 'TKoV3')) +
    geom_text(aes(label = number_of_genes, y = number_of_genes / 2), size = 15) +
    theme(axis.text.x = element_text(size = 50, colour = 'black'),
        axis.text.y = element_text(size = 50),
        axis.title.y = element_text(size = 50, vjust=0.5, margin = margin(r = 20)),
        legend.title = element_text(size = 50),
        legend.text = element_text(size = 50),
        line = element_line((size = 2), colour = 'black'),
        plot.margin = margin(1, 1, 1, 1)) +
    scale_fill_manual(values = custom_colors,
                      name = 'Gene symbol in sgRNA ID',
                      labels = c('match' = 'Approved symbol', 
                                 'mismatch' = 'Previous/alias symbol')) +
    theme(legend.key.size = unit(1.5, "cm")) +
    scale_y_continuous(limits = c(0, 20000), breaks = seq(0, 20000, by = 5000))


barplot

ggsave(
    path = './',
    width = 25,
    height = 15,
    dpi = 300,
    plot = barplot,
    filename = paste('histogram_exon_annotation_after.png', sep = ''))


```

# Original sgRNA library vs Refined sgRNA library percentage

```{r}
# import sample sgRNA library

import_sgrna_library <- function(data_dir) {
  library <- read_tsv(data_dir, col_names = FALSE)
  colnames(library) <- c('sgRNA', 'spacer', 'gene')
  return(library)
}

brunello <- import_sgrna_library('./public_crispr_library/broadgpp-brunello-library-contents.tsv')
toronto <- import_sgrna_library('./public_crispr_library/tkov3_guide_sequence.tsv')
cellecta_M1 <- import_sgrna_library('./cellecta_library/sample_M1_150K.tsv')
cellecta_M2 <- import_sgrna_library('./cellecta_library/sample_M2_150K.tsv')
cellecta_M3 <- import_sgrna_library('./cellecta_library/sample_M3_150K.tsv')

# combined cellecta and remove duplicates
cellecta <- rbind(cellecta_M1, cellecta_M2, cellecta_M3)
cellecta <- cellecta %>%
  arrange(sgRNA) %>%
  distinct()

refined_brunello <- import_sgrna_library('./output_cleaning/broadgpp-brunello-library-contents_65K_refined.tsv')
refined_toronto <- import_sgrna_library('./output_cleaning/tkov3_guide_sequence_62K_refined.tsv')
refined_cellecta_M1 <- import_sgrna_library('./output_cleaning/sample_M1_150K_48K_refined.tsv')
refined_cellecta_M2 <- import_sgrna_library('./output_cleaning/sample_M2_150K_45K_refined.tsv')
refined_cellecta_M3 <- import_sgrna_library('./output_cleaning/sample_M3_150K_42K_refined.tsv')

refined_cellecta <- rbind(refined_cellecta_M1, refined_cellecta_M2, refined_cellecta_M3)
refined_cellecta <- refined_cellecta %>%
  arrange(sgRNA) %>%
  distinct()

percentage_sgrna <- function(original_library, refined_library) {
    percentage = round(n_distinct(refined_library$sgRNA) / n_distinct(original_library$sgRNA) * 100, 2)
    return(percentage)
}

percentage_sgrna_brunello <- percentage_sgrna(brunello, refined_brunello)
percentage_sgrna_toronto <- percentage_sgrna(toronto, refined_toronto)
percentage_sgrna_cellecta <- percentage_sgrna(cellecta, refined_cellecta)

percentage_sgrna_removed_brunello <- 100 - percentage_sgrna_brunello
percentage_sgrna_removed_toronto <- 100 - percentage_sgrna_toronto
percentage_sgrna_removed_cellecta <- 100 - percentage_sgrna_cellecta

library <- c('brunello', 'toronto', 'cellecta')

# Create a data frame for plotting with both percentages
percentage_existing <- c(percentage_sgrna_brunello, percentage_sgrna_toronto, percentage_sgrna_cellecta)
percentage_removed <- c(percentage_sgrna_removed_brunello, percentage_sgrna_removed_toronto, percentage_sgrna_removed_cellecta)
percentage_data <- data.frame(library, percentage_removed, percentage_existing)

# Reshape data for stacked bar chart
percentage_data_long <- percentage_data %>%
    pivot_longer(cols = c("percentage_removed", "percentage_existing"), 
                 names_to = "status", 
                 values_to = "percentage")

custom_colors <- c("percentage_removed" = "mistyrose", "percentage_existing" = "lightgreen")

stacked_barplot <- ggplot(percentage_data_long, aes(x = factor(library, levels = c('cellecta', 'brunello', 'toronto')), y = percentage, fill = status)) +
    geom_bar(stat = "identity", width = 0.7) +
    theme_minimal() +
    labs(title = "", x = "", y = "Percentage of sgRNAs (%)") +
    scale_x_discrete(labels = c('brunello' = 'Brunello \n 77K sgRNAs',
                                'cellecta' = 'Cellecta \n 150K sgRNAs',
                                'toronto' = 'Toronto \nKnockout V3 \n 71K sgRNAs')) +
    geom_text(aes(label = paste0(percentage, "%")), position = position_stack(vjust = 0.5), size = 10) +
    theme(axis.text.x = element_text(size = 23),
          axis.text.y = element_text(size = 30),
          axis.title.y = element_text(size = 30, vjust = 0.5, margin = margin(r = 20)),
          legend.title = element_text(size = 25),
          legend.text = element_text(size = 25),
          plot.margin = margin(1, 1, 1, 1)) +
    scale_fill_manual(values = custom_colors,
                      name = 'sgRNA Status',
                      labels = c('percentage_existing' = 'Retained sgRNAs', 
                                 'percentage_removed' = 'Eliminated sgRNAs')) +
    theme(legend.key.size = unit(1.5, "cm"))

stacked_barplot

ggsave(
    path = './',
    width = 15,
    height = 15,
    dpi = 300,
    plot = stacked_barplot,
    filename = paste('stacked_barplot_sgRNA.png', sep = ''))


```


```{r}
library(tidyverse)

# library <- c('brunello', 'brunello', 'toronto', 'toronto', 'cellecta', 'cellecta')
# number_of_genes <- c(16967, 1056, 16313, 1125, 17042, 1376)
# annotation <- c('match', 'mismatch', 'match', 'mismatch', 'match', 'mismatch')
# library <- c('brunello', 'toronto', 'cellecta')
# number_of_genes <- c(17927, 17419, 17959)
# annotation <- c('match', 'match', 'match')

# Example data
df <- tribble(
  ~library,     ~condition, ~status,     ~number_of_genes,
  "cellecta",  "Before",   "match",     17042,
  "cellecta",  "Before",   "mismatch",    1376,
  "cellecta",  "After",    "match",    17959,
  "brunello",  "Before",   "match",     16967,
  "brunello",  "Before",   "mismatch",    1056,
  "brunello",  "After",    "match",    17927,
  "toronto_v3",  "Before",   "match",     16313,
  "toronto_v3",  "Before",   "mismatch",    1125,
  "toronto_v3",  "After",    "match",    17419
)

# Ensure 'condition' is ordered so Before always comes first
df <- df %>%
  mutate(
    condition = factor(condition, levels = c("Before", "After")),
    status = factor(status, levels = c("mismatch", "match"))  # order matters here
  )

library <- c("brunello", "toronto_v3", "cellecta")
factor <- c(2,3,1)

# Plot
stacked_barplot_correction <- ggplot(df, aes(x = reorder(library, -number_of_genes), y = number_of_genes, fill = status)) +
geom_col() +
  facet_grid(~ condition, 
             scales = "free_x",
             space = "free_x",
             switch = "x",
             labeller = as_labeller(c("Before" = "Before gene symbol update", "After" = "After gene symbol update"))) +
  scale_fill_manual(values = c('match' = 'lightgreen', 'mismatch' = 'mistyrose'),
                      name = 'Gene symbol in sgRNA ID',
                      labels = c('match' = 'Approved symbol', 
                                 'mismatch' = 'Previous/alias symbol')) +
  labs(
    title = "",
    x = "",
    y = "Number of Genes",
    fill = "Gene Status"
  ) +
    scale_x_discrete(labels = c('brunello' = 'Brunello',
                                'cellecta' = 'Cellecta',
                                'toronto_v3' = 'TKoV3')) +
  theme_bw(base_line_size = 15) +
  theme(legend.position = 'right',
        axis.text.x = element_text(size = 50, colour = 'black'),
        axis.text.y = element_text(size = 50),
        axis.title.y = element_text(size = 50, vjust=0.5, margin = margin(r = 20)),
        legend.title = element_text(size = 50),
        legend.text = element_text(size = 50),
        line = element_line((size = 4), colour = 'black'),
        strip.text = element_text(size = 50),
        strip.placement = "outside",
        strip.background = element_blank(),
        strip.text.x = element_text(vjust = 0.5),
        plot.margin = margin(1, 1, 1, 1)) +
  # Optional: Label bars as "Before"/"After"
  
  geom_text(
    data = df,
    (aes(label = number_of_genes)),
    stat = "identity",
    position = position_stack(vjust = 0.5),
    size = 15,
    color = "black"
  ) +
    scale_y_continuous(limits = c(0, 20000), breaks = seq(0, 20000, by = 5000))

ggsave(
    path = './',
    width = 35,
    height = 15,
    dpi = 300,
    plot = stacked_barplot_correction,
    filename = paste('stacked_barplot_correction.png', sep = ''))


```