---
title: "GuideRefine: a tool to remove multi-target-/off-target sgRNAs from genome-wide CRISPR-KO screen libraries"
subtitle: "Developed with R v4.5.2"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
    html_document:
      theme: cerulean
      toc: true
      toc_float: true
      number_sections: true
      self_contained: true
      
params:
  output_filename: "output.html"
  # library directory and control guides name
  library_dir: "./public_crispr_library/"
  sgrna_library: "Jacquere_PerGuideAnnotations_Quota4"
  terms: ["CONTROL", "Control", "control", "INTRON", "Intron", "intron", "NO_SITE", "ONE_SITE_INTERGENIC"]
  # reference genome and hgnc annotation
  ref_bsgenome: "BSgenome.Hsapiens.NCBI.T2TCHM13v2.0"
  ccds_filename_dir: "annotation_file/T2T-CHM13v2.0_gene_annot_granges.rds"
  hgnc_link: "https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt"
  # alignment file location and criteria
  guide_aln_granges_genome: "T2T-CHM13v2.0"
  aln_file_dir: "./object_intermediate/T2T-CHM13_bsgenome/"
  check_index_file: "T2T-CHM13_BSgenome.1.ebwt"
  bowtie_index_dir: "bowtie_index/T2T-CHM13_bsgenome/"
  name_prefix_for_index: "T2T-CHM13_BSgenome"
  fasta_file: ""  # path to reference FASTA file; required only when bowtie index has not been built yet
  remove_all_single_mismatch: TRUE
  # output location
  output_dir: "./output_cleaning/"
  
---

# Input data and Methods

## Input data
1. sgRNA library TSV file input for MAGeCK (Format: sgRNA, spacer, gene; without header)
2. Reference genome indexed using bowtie tool (can be hg38 or T2T-CHM13)
3. CCDS annotation file. Look for the current release here: https://ftp.ncbi.nlm.nih.gov/pub/CCDS/current_human/
4. HGNC (Hugo Gene Nomenclature Committee) annotation file. Look for the current release here: https://www.genenames.org/download/custom/

The current HGNC dataset were retrieved from https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt

## Methods

1.  Import required packages
2.  Define any filename
3.  Import sgRNA library and create dir for output (output_cleaning) and intermediate files (object_intermediate)
4.  Alignment of sgRNA to reference genome (hg38 or T2T-CHM13) using Bowtie via crisprBowtie (max 2 mismatches)
5.  Post alignment data processing
6.  Find and drop multi-target sgRNAs
7.  Find and drop single mismatch sgRNAs (optional)
8.  Find and drop sgRNA align with single mismatch at the PAM-distal
9.  Find and drop sgRNA align with double mismatch at the PAM-distal
10. CCDS annotation of sgRNA library
11. HGNC annotation of sgRNA library
12. Clean sgRNA library
13. Discard any genes targeted by less than 3 sgRNAs
14. Calculate how many sgRNAs or genes lost
15. Export the refined sgRNA library TSV file and excel file report

# Pipeline

```{r Import required library, include = FALSE}

source("GuideRefine_functions.R")

library(crisprBase)
library(crisprBowtie)
library(GenomeInfoDb)
library(params$ref_bsgenome, character.only = TRUE)
library(GenomicRanges)
library(stringr)
library(directlabels)
library(tidyverse)
library(knitr)
library(openxlsx)

```

## Parameters for the pipeline

```{r User defined variable (Input in params), echo = FALSE}

# main sgRNA library file location
file_dir_location <- params$library_dir

# name of sgRNA library (without file format)
sgRNA_library_selected <- params$sgrna_library

# detect any control/intron control sgRNA from your library
terms <- params$terms
terms_for_filtering <- paste(terms, collapse = "|")

# change your reference genome here
ref_bsgenome_selected <- get(params$ref_bsgenome)

# directory for NCBI CCDS data or .RDS file processed from T2T_gff_to_ccds_conversion.Rmd
ccds_filename_dir <- params$ccds_filename_dir

# guide align granges genome (hg38 or T2T-CHM13v2.0)
guide_aln_granges_genome <- params$guide_aln_granges_genome

# name output for sgRNA library alignment (.CSV file)
aln_file_dir <- params$aln_file_dir
aln_2_mismatch_filename <- paste(sgRNA_library_selected, "_aln.csv", sep ="")

# Bowtie alignment indexing
check_index_file <- params$check_index_file
bowtie_index_dir <- params$bowtie_index_dir
name_prefix_for_index <- params$name_prefix_for_index
fasta_file <- params$fasta_file

# alignment criteria
remove_all_single_mismatch <- params$remove_all_single_mismatch

paste("sgRNA library name:", sgRNA_library_selected)
paste("CCDS or .RDS file:", ccds_filename_dir)
paste("Genome version:", unique(genome(ref_bsgenome_selected)))

```

```{r Create folder and import library, include = FALSE}
dir.create("object_intermediate", recursive = TRUE, showWarnings = FALSE)
dir.create(aln_file_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
sgRNA_library <- read_tsv(paste(file_dir_location, sgRNA_library_selected, ".tsv", sep = ""), col_names = FALSE)
colnames(sgRNA_library) <- c("sgRNA", "spacer", "gene")

```

```{r Check for HG38 index files, include = FALSE}

list.files(path = bowtie_index_dir, pattern = "*.ebwt$")
file_test("-f", paste(bowtie_index_dir, check_index_file, sep="")) # returns TRUE

```

```{r Indexing HG38 if indexing have not been done previously, include = FALSE}
if (file_test("-f", paste(bowtie_index_dir, check_index_file, sep="")) == FALSE) {
  Rbowtie::bowtie_build(fasta_file,
                         outdir = bowtie_index_dir,
                         force = TRUE,
                         prefix = name_prefix_for_index)
} else {
  print(paste(check_index_file, "is in bowtie_index directory, indicating reference genome indexing by bowtie have been done previously"))
}
```

```{r Check whether alignment has been done, include = FALSE}

if (file_test("-f", paste(aln_file_dir, aln_2_mismatch_filename, sep = "")) == FALSE) {
      data(SpCas9, package="crisprBase")
      crisprNuclease <- SpCas9
      
      spacers_sequence <- c(unique(sgRNA_library$spacer))
      
      bowtie_index <- file.path(bowtie_index_dir, name_prefix_for_index)
      
      # change your reference genome here
      # seqlevelsStyle(genome) <- "UCSC"
      
      alignment_guides <- runCrisprBowtie(spacers_sequence,
                                           mode = "spacer",
                                           bsgenome = ref_bsgenome_selected,
                                           crisprNuclease = crisprNuclease,
                                           n_mismatches = 2,
                                           force_spacer_length = TRUE,
                                           canonical = TRUE,
                                           bowtie_index = bowtie_index)
      
      alignment_data <- sgRNA_library %>%
        left_join(alignment_guides, by = join_by(spacer))
      
      write.csv(alignment_data, 
                file = paste(aln_file_dir, aln_2_mismatch_filename, sep = ""), 
                row.names = FALSE)
  
} else {
  paste(aln_2_mismatch_filename, "file is in object_intermediate folder, this indicates the alignment with max 2 mismatches has been conducted previously")
}

```

```{r import csv file from the alignment, echo = FALSE, include = FALSE}

library_alignment <- read_csv(paste(aln_file_dir, aln_2_mismatch_filename, sep = ""))

```

```{r, echo = FALSE, include = FALSE}

track_table <- data.frame(matrix(ncol = 4, nrow = 0))
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = "count beginning")
```

```{r Select original gene list and remove any alignment to non standard chromosome, include = FALSE}
gene_original <- c(unique(library_alignment$gene))

# different notation between NCBI and UCSC genome format (chr and number)`

# genome(ref_bsgenome_selected)
# seqlevelsStyle(ref_bsgenome_selected)

seq_style <- seqlevelsStyle(ref_bsgenome_selected)

if ("NCBI" %in% seq_style) {

  list_chromosome <- c(1:22, "X", "Y")
  library_alignment <- library_alignment %>%
    filter(chr %in% list_chromosome | is.na(chr))

} else if ("UCSC" %in% seq_style) {

  list_chromosome <- paste0("chr", c(1:22, "X", "Y"))
  library_alignment <- library_alignment %>%
    filter(chr %in% list_chromosome | is.na(chr))

}

```

## Report of multi-target sgRNAs

```{r Find multi-target sgRNAs, include = FALSE}

# detect any multi-target sgRNAs
result_multi_target_sgrna <- multi_target_sgrna(library_alignment)

# remove multi-target sgRNAs
library_alignment <- library_alignment %>%
  filter(!sgRNA %in% result_multi_target_sgrna$list_guides)

# genes lost due to multi-target sgRNAs
genes_lost_multi_target_sgrna <- setdiff(gene_original, unique(library_alignment$gene))

# differences between gene original and gene lost
gene_original <- setdiff(gene_original, genes_lost_multi_target_sgrna)

```

```{r Report of multi-target sgRNAs, echo = FALSE}

# number of gene lost due to multi-target sgRNAs
print(paste("Number of multi-target sgRNAs: ", length(result_multi_target_sgrna$list_guides), sep = ""))
print(paste("Number of genes lost due to multi-target sgRNAs: ", length(genes_lost_multi_target_sgrna), sep = ""))

```

```{r, include = FALSE}
# update track table
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = paste(length(result_multi_target_sgrna$list_guides), " multi-target guides discarded"))
```

## Report of single mismatch sgRNAs

```{r Find sgRNA aligned with single mismatch (optional), include = FALSE}

if(remove_all_single_mismatch == TRUE) {
  
  # detect any single mismatch sgRNA
  result_single_mismatch_sgrna <- single_mismatch_sgrna(library_alignment)

  # remove single mismatch sgRNAs
  library_alignment <- library_alignment %>%
    filter(!sgRNA %in% result_single_mismatch_sgrna$list_guides)
  
  # genes lost due to single mismatch sgRNA
  genes_lost_single_mismatch_sgrna <- setdiff(gene_original, unique(library_alignment$gene))
  
  # differences between gene original and genes lost
  gene_original <- setdiff(gene_original, genes_lost_single_mismatch_sgrna)
  
  # update track table
  track_table <- function_df_count(track_table,
                                   count_sgrna = length(unique(library_alignment$sgRNA)),
                                   num_genes = length(unique(library_alignment$gene)),
                                   num_alignments = length(library_alignment$sgRNA),
                                   notes = paste(length(result_single_mismatch_sgrna$list_guides), " single mismatch sgRNA discarded"))
  
  }

```

```{r Report of single mismatch sgRNAs, echo = FALSE}

if (remove_all_single_mismatch == TRUE) {
  # number of genes lost due to single mismatch sgRNAs
  print(paste("Number of single mismatch sgRNAs: ", length(result_single_mismatch_sgrna$list_guides), sep = ""))
  print(paste("Number of genes lost due to single mismatch sgRNAs: ", length(genes_lost_single_mismatch_sgrna), sep = ""))
}

```

## Report of PAM-distal mismatch sgRNAs

```{r Find sgRNAs with mismatch at PAM distal, include = FALSE}

# detect any pam-distal mismatch sgRNA
result_pam_distal_mismatch <- pam_distal_mismatch(library_alignment)

# ---------------------------------------------------------------------------------------------------------------------------------------------------------

# remove pam-distal single mismatch sgRNA
library_alignment <- result_pam_distal_mismatch$pam_distal_df %>%
  filter(!sgRNA %in% result_pam_distal_mismatch$list_guides_pam_single)

# genes lost due to pam-distal single mismatch sgRNA
genes_lost_pam_distal_single_mismatch_sgrna <- setdiff(gene_original, unique(library_alignment$gene))

# differences between gene original and genes lost
gene_original <- setdiff(gene_original, genes_lost_pam_distal_single_mismatch_sgrna)

# update track table
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = paste(length(result_pam_distal_mismatch$list_guides_pam_single), " PAM-distal single mismatch sgRNA discarded"))

# ---------------------------------------------------------------------------------------------------------------------------------------------------------

# remove pam-distal double mismatch sgRNA
library_alignment <- library_alignment %>%
  filter(!sgRNA %in% result_pam_distal_mismatch$list_guides_pam_double)

# genes lost due to pam-distal double mismatch sgRNA
genes_lost_pam_distal_double_mismatch_sgrna <- setdiff(gene_original, unique(library_alignment$gene))

# differences between gene original and genes lost
gene_original <- setdiff(gene_original, genes_lost_pam_distal_double_mismatch_sgrna)

# update track table
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = paste(length(result_pam_distal_mismatch$list_guides_pam_double), " PAM-distal double mismatch sgRNA discarded"))

```

```{r Report of sgRNA with PAM-distal mismatch, echo = FALSE}

# number of genes lost due to PAM-distal single mismatch sgRNAs
print(paste("Number of PAM-distal single mismatch sgRNAs: ", length(result_pam_distal_mismatch$list_guides_pam_single), sep = ""))
print(paste("Number of genes lost due to PAM-distal single mismatch sgRNAs: ", length(genes_lost_pam_distal_single_mismatch_sgrna), sep = ""))

# number of genes lost due to PAM-distal double mismatch sgRNAs
print(paste("Number of PAM-distal double mismatch sgRNAs: ", length(result_pam_distal_mismatch$list_guides_pam_double), sep = ""))
print(paste("Number of genes lost due to PAM-distal double mismatch sgRNAs: ", length(genes_lost_pam_distal_double_mismatch_sgrna), sep = ""))

```

```{r Separate control or intron control sgRNA with the rest of sgRNA, include = FALSE}

# separate control sgRNA from the rest of sgRNA
sgRNA_library_control <- library_alignment %>%
  filter(str_detect(gene, terms_for_filtering)) %>%
  select(sgRNA, spacer, gene) %>%
  distinct()

# separate sgRNA not targeting anywhere (by checking sgRNA that does not have protospacer)
sgRNA_not_target_anywhere_ids <- library_alignment %>%
  filter(is.na(protospacer) & !str_detect(gene, terms_for_filtering)) %>%
  pull(sgRNA) %>%
  unique()

sgRNA_not_target_anywhere <- library_alignment %>%
  filter(is.na(protospacer) & !str_detect(gene, terms_for_filtering)) %>%
  select(sgRNA, gene) %>%
  distinct() %>%
  count(gene) %>%
  dplyr::rename("sgRNA not target anywhere" = "n")

# filter gene not in control and drop sgRNAs not targeting anywhere
library_alignment <- library_alignment %>%
  filter(!gene %in% unique(sgRNA_library_control$gene)) %>%
  drop_na()

# detected control sgRNA
detected_control_sgrna <- unique(sgRNA_library_control$gene)

```

```{r, echo = FALSE}
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = paste0(sum(sgRNA_not_target_anywhere$`sgRNA not target anywhere`), " sgRNAs not target anywhere discarded"))
```

```{r Using NCBI CCDS data to find the overlap of sgRNA alignment with exon annotation, include = FALSE}

# if input of CCDS is .txt file, then use this branch
if (grepl("\\.txt$", ccds_filename_dir)) {
  
    print(paste("Processing:", ccds_filename_dir))
  
    # import ccds data
    ccds <- read_ccds_data(ccds_filename = ccds_filename_dir)
    
    # transform gene annotation data
    transformed_gene_annotation <- transform_gene_annotation_ccds(ccds, chosen_genome = guide_aln_granges_genome)
    
    # take ccds_exon data from transformed gene annotation
    ccds_exon <- transformed_gene_annotation$ccds_exon_df
    
    # take granges data from transformed gene annotation
    gene_annot_granges <- transformed_gene_annotation$gene_annot_granges_df
   
    
} else if (grepl("\\.rds$", ccds_filename_dir, ignore.case = TRUE)) {
  
    print(paste("Processing:", ccds_filename_dir))
  
    # import .rds data
    gene_annot_granges <- readRDS(file = ccds_filename_dir)
  
}

# add alignment id for every sgRNA alignment
library_alignment <- library_alignment %>% 
  mutate(unique_aln_id = paste0("sgrna_aln_", row_number())) %>%
  relocate(unique_aln_id, .before = sgRNA)

# be careful switching between GRCh38 and T2T-CHM13v2.0

# make granges from alignment data
guide_aln_granges <- make_granges_from_alignment_data(library_alignment, 
                                                      ref_bsgenome_selected = ref_bsgenome_selected)

# final result of gene_df in a dataframe containing the overlap of gene annotation and alignment
gene_df <- find_overlaps_gene_annotation_and_alignment(guide_aln_granges = guide_aln_granges,
                                                       gene_annot_granges = gene_annot_granges)

```

## Validation of gene symbol for the alignment and annotation data

**Alignment validation to check whether there is a symbol mismatch between the gene symbol and the exon annotation**

```{r Validation of gene symbol for the alignment and annotation data - Strictly pick 0 mismatch alignment, echo = FALSE}

visualize_gene_symbol(gene_df)

```

**Correct the sgRNA according to targeted gene symbol**

```{r Correct the sgRNA according to targeted gene symbol, echo = FALSE}

# update gene symbol based on the sgRNA alignment location to exon
gene_df_corrected <- update_gene_symbol(gene_df)

# Histogram of corrected gene symbol
visualize_gene_symbol(gene_df_corrected[["gene_symbol_corrected"]])

```

```{r Find genes lost after annotation, include = FALSE}

# gene_track_annotation : a dataframe of the old and approved gene symbol
gene_track_annotation <- gene_df_corrected[["gene_symbol_check"]] %>%
  filter(notes == "symbol mismatch") %>%
  select(sgrna, gene, notes) %>%
  mutate(old_gene = str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)")) %>%
  relocate(old_gene, .before = gene) %>%
  distinct() %>%
  select(-sgrna) %>%
  group_by(old_gene, gene, notes) %>%
  summarise(`Additional sgRNA (Corrected)` = n()) %>%
  ungroup()

# corrected_sgrna_report : a dataframe from gene_track_annotation used for the report
corrected_sgrna_report <- gene_track_annotation %>%
  select(-c(old_gene, notes))

prev_gene <- c(gene_track_annotation$old_gene)
align_gene <- c(gene_track_annotation$gene)

gene_original_corrected <- sapply(gene_original, function(gene_query){
  match_index <- match(gene_query, prev_gene)
  if (!is.na(match_index)) {
   align_gene[match_index]
  } else {
    gene_query
  }
})

gene_original_corrected <- as.character(gene_original_corrected)

paste("Corrected gene name: ")
setdiff(gene_original_corrected, gene_original)

exclude_control <- c(unique(gene_df_corrected[["gene_symbol_corrected"]]$gene), detected_control_sgrna)

genes_lost_after_annotation <- setdiff(gene_original_corrected, exclude_control)
gene_original_corrected <- setdiff(gene_original_corrected, genes_lost_after_annotation)

```

[**Genes lost after annotation**]{style="color:red"}

**Genes lost at this stage means that its sgRNA was not aligned to anywhere in the genome**

```{r Genes lost after gene annotation, echo = FALSE}

print(paste("number of genes lost following the annotation: ", length(genes_lost_after_annotation), sep = ""))

```

```{r, include = FALSE}
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(gene_df_corrected[["gene_symbol_corrected"]]$sgrna)),
                                 num_genes = length(unique(gene_df_corrected[["gene_symbol_corrected"]]$gene)),
                                 num_alignments = length(gene_df_corrected[["gene_symbol_corrected"]]$sgrna),
                                 notes = paste(length(unique(gene_df_corrected[["gene_symbol_corrected"]]$sgrna)), " sgRNA targeting exon only"))
```

```{r Find sgRNAs failed to target exon, include = FALSE}

list_sgRNA_not_target_exon <- setdiff(unique(library_alignment$sgRNA), unique(gene_df_corrected[["gene_symbol_corrected"]]$sgrna))

sgRNA_not_target_exon <- library_alignment %>%
  select(sgRNA, gene) %>%
  filter(sgRNA %in% list_sgRNA_not_target_exon) %>%
  distinct() %>%
  count(gene) %>%
  filter(!gene %in% gene_track_annotation$old_gene) %>%
  dplyr::rename("sgRNA not target exon" = "n")

```

```{r Add HGNC annotation of sgRNA library, include = FALSE}

hgnc_data <- read_tsv(params$hgnc_link)

hgnc_selected <- hgnc_data %>%
  select(symbol, name, locus_group, locus_type) %>%
  dplyr::rename("gene" = "symbol")

sgRNA_library_gene_df <- gene_df_corrected[["gene_symbol_corrected"]] %>%
  left_join(hgnc_selected, by = "gene")

print(paste("count sgRNAs: ", length(unique(sgRNA_library_gene_df$sgrna))))
print(paste("alignments left after data joining with hgnc:", length(sgRNA_library_gene_df$sgrna)))

```

```{r Filter sgRNAs targeting protein-coding genes only, include = FALSE}

sgRNA_library_gene_df_annot <- sgRNA_library_gene_df %>%
  filter(locus_group == "protein-coding gene")

print(paste("count sgRNAs: ", length(unique(sgRNA_library_gene_df_annot$sgrna))))
print(paste("alignments left after data selection of protein-coding gene only:", length(sgRNA_library_gene_df_annot$sgrna)))

```

```{r Find genes failed in the annotation, include = FALSE}

exclude_control_2 = c(unique(sgRNA_library_gene_df_annot$gene), detected_control_sgrna)
genes_not_pcg <- setdiff(sgRNA_library_gene_df$gene, exclude_control_2)

genes_not_pcg_df <- sgRNA_library_gene_df %>%
  filter(gene %in% genes_not_pcg)
  
```

[**sgRNAs not targeting protein coding genes**]{style="color:red"}

```{r Selected only protein-coding genes, echo = FALSE}

print(paste("number of non protein-coding genes: ", length(genes_not_pcg), sep = ""))

```

```{r, include = FALSE}
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(sgRNA_library_gene_df_annot$sgrna)),
                                 num_genes = length(unique(sgRNA_library_gene_df_annot$gene)),
                                 num_alignments = length(sgRNA_library_gene_df_annot$sgrna),
                                 notes = paste(length(unique(sgRNA_library_gene_df_annot$sgrna)), " sgRNA successfully annotated"))

```

```{r Obtain refined sgRNA library, echo = FALSE, include = FALSE}

sgRNA_library_cleaned <- sgRNA_library_gene_df_annot %>%
  select(sgrna, spacer, gene) %>%
  distinct() %>%
  dplyr::rename("sgRNA" = "sgrna")

```

```{r Discard any genes targeted by < 3 sgRNAs, echo = FALSE, include = FALSE}

guides_below_3 <- sgRNA_library_cleaned %>%
  count(gene) %>%
  filter(n < 3)

sgRNA_library_cleaned <- sgRNA_library_cleaned %>%
  filter(!gene %in% guides_below_3$gene)

count_guides_per_gene <- sgRNA_library_cleaned %>%
  count(gene) %>%
  count(n) %>%
  mutate(`total number of guides` = n * nn) %>%
  dplyr::rename("number of guides per gene" = "n") %>%
  dplyr::rename("how many genes" = "nn")

num_original_genes <- length(unique(sgRNA_library$gene))
num_genes_after_cleaning_3_guides <- length(unique(sgRNA_library_cleaned$gene))

paste("total number of sgRNA: ", sum(count_guides_per_gene$`total number of guides`))
paste("number of original target genes in the library: ", length(unique(sgRNA_library$gene)))
paste("number of genes after exclude gene targeted by guides < 3: ", length(unique(sgRNA_library_cleaned$gene)))
paste("number of genes lost: ", num_original_genes-num_genes_after_cleaning_3_guides)

```

## Final report of refined sgRNA library

```{r Combine the control sgRNA back with the original dataframe, include = FALSE}	

# Add control sgRNA (sgRNA_library_control)
sgRNA_library_refined <- rbind(sgRNA_library_control, sgRNA_library_cleaned)

```

```{r, echo = FALSE}

track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(sgRNA_library_refined$sgRNA)),
                                 num_genes = length(unique(sgRNA_library_refined$gene)),
                                 num_alignments = "-",
                                 notes = paste(length(unique(guides_below_3$gene)), " genes removed (fewer than 3 sgRNAs)"))

colnames(track_table) <- c("count_sgrna", "count_genes", "count_alignments", "notes")
kable(track_table)

```

```{r Calculate total sgRNAs and genes lost, include = FALSE}

total_sgRNA_or_gene_lost <- data.frame("gene" = sgRNA_library$gene) %>%
  count(gene) %>%
  dplyr::rename("sgRNA number" = "n") %>%
  left_join(result_multi_target_sgrna$dropped_genes_num, by = "gene") %>%
  mutate(`multi-target sgRNA` = replace_na(`multi-target sgRNA`, 0)) %>%
  
  {
    tmp <- .
    if (remove_all_single_mismatch == TRUE) {
      tmp <- tmp %>%
        left_join(result_single_mismatch_sgrna$dropped_genes_num, by = "gene") %>%
        mutate(`single mismatch sgRNA` = replace_na(`single mismatch sgRNA`, 0))
    }
    tmp
  } %>%
  
  left_join(result_pam_distal_mismatch$dropped_genes_num_pam_single, by = "gene") %>%
  mutate(`PAM distal single mismatch` = replace_na(`PAM distal single mismatch`, 0)) %>%
  left_join(result_pam_distal_mismatch$dropped_genes_num_pam_double, by = "gene") %>%
  mutate(`PAM distal double mismatch` = replace_na(`PAM distal double mismatch`, 0)) %>%
  
  left_join(sgRNA_not_target_anywhere, by = "gene") %>%
  mutate(`sgRNA not target anywhere` = replace_na(`sgRNA not target anywhere`, 0)) %>%
  left_join(sgRNA_not_target_exon, by = "gene") %>%
  mutate(`sgRNA not target exon` = replace_na(`sgRNA not target exon`, 0)) %>%
  
  full_join(corrected_sgrna_report, by = "gene") %>%
  mutate(`Additional sgRNA (Corrected)` = replace_na(`Additional sgRNA (Corrected)`, 0)) %>%
  replace(is.na(.), 0) %>%
  {
    tmp <- .
    if (remove_all_single_mismatch == TRUE) {
      tmp <- tmp %>%
        mutate(`counted total sgRNA` = `sgRNA number` - (`multi-target sgRNA` +
                                                         `single mismatch sgRNA` +
                                                         `PAM distal double mismatch` +
                                                         `sgRNA not target anywhere` +
                                                         `sgRNA not target exon`) +
                                                         `Additional sgRNA (Corrected)`)
    } else {
      tmp <- tmp %>%
        mutate(`counted total sgRNA` = `sgRNA number` - (`multi-target sgRNA` +
                                                         `PAM distal single mismatch` +
                                                         `PAM distal double mismatch` +
                                                         `sgRNA not target anywhere` +
                                                         `sgRNA not target exon`) +
                                                         `Additional sgRNA (Corrected)`)
    }
    tmp
  } %>%
  
  mutate(`symbol change` = case_when(gene %in% gene_track_annotation$old_gene ~
                                       gene_track_annotation$gene[match(gene, gene_track_annotation$old_gene)],
                                        TRUE ~ "no")) %>%
  mutate(`protein-coding gene` = case_when(str_detect(gene, terms_for_filtering) ~ "no",
                                           gene %in% genes_not_pcg ~ "no",
                                           TRUE ~ "yes"))

crosscheck_count <- sgRNA_library_refined %>%
  count(gene) %>%
  dplyr::rename("actual total sgRNA" = "n")

total_sgRNA_or_gene <- total_sgRNA_or_gene_lost %>%
  left_join(crosscheck_count, by = join_by(gene)) %>%
  mutate(`actual total sgRNA` = replace_na(`actual total sgRNA`, 0)) %>%
  relocate(`actual total sgRNA`, .after = `counted total sgRNA`) %>%
  mutate(`number match` = mapply(identical,                 
                                 as.numeric(`counted total sgRNA`), 
                                 as.numeric(`actual total sgRNA`))) %>%
  relocate(`number match`, .after = `actual total sgRNA`) %>%
  mutate(`quality check` = case_when((`symbol change` != "no") ~ "not counted (symbol change)",
                                     (`actual total sgRNA` > 2 & `protein-coding gene` == "yes") ~ "pass",
                                     TRUE ~ "fail")) %>%
  mutate(`quality check` = ifelse(gene %in% unique(sgRNA_library_control$gene), "pass", `quality check`)) %>%
  arrange(factor(gene, levels = c(unique(sgRNA_library_control$gene), setdiff(unique(gene), unique(sgRNA_library_control$gene))))) %>%
  distinct()

```

```{r Create a dataframe what sgRNA lost during each process, include = FALSE}

if (remove_all_single_mismatch == TRUE) {
  max_len <- max(length(result_multi_target_sgrna$list_guides),
                 length(result_single_mismatch_sgrna$list_guides),
                 length(result_pam_distal_mismatch$list_guides_pam_single),
                 length(result_pam_distal_mismatch$list_guides_pam_double),
                 length(sgRNA_not_target_anywhere_ids),
                 length(list_sgRNA_not_target_exon))
} else {
  max_len <- max(length(result_multi_target_sgrna$list_guides),
                 length(result_pam_distal_mismatch$list_guides_pam_single),
                 length(result_pam_distal_mismatch$list_guides_pam_double),
                 length(sgRNA_not_target_anywhere_ids),
                 length(list_sgRNA_not_target_exon))
}

pad <- function(x, n) {
  length(x) <- n 
  return(x)
}


if (remove_all_single_mismatch == TRUE) {
  list_discarded_sgrna <- data.frame(
    multi_target_guides = pad(result_multi_target_sgrna$list_guides, max_len),
    single_mismatch_guides = pad(result_single_mismatch_sgrna$list_guides, max_len),
    pam_distal_single_mismatch_guides = pad(result_pam_distal_mismatch$list_guides_pam_single, max_len),
    pam_distal_double_mismatch_guides = pad(result_pam_distal_mismatch$list_guides_pam_double, max_len),
    guides_not_target_anywhere = pad(sgRNA_not_target_anywhere_ids, max_len),
    stringsAsFactors = FALSE)
} else {
  list_discarded_sgrna <- data.frame(
    multi_target_guides = pad(result_multi_target_sgrna$list_guides, max_len),
    pam_distal_single_mismatch_guides = pad(result_pam_distal_mismatch$list_guides_pam_single, max_len),
    pam_distal_double_mismatch_guides = pad(result_pam_distal_mismatch$list_guides_pam_double, max_len),
    guides_not_target_anywhere = pad(sgRNA_not_target_anywhere_ids, max_len),
    stringsAsFactors = FALSE)
  }

```

## Export the refined library, sgRNA cleaning report, and list discarded sgRNA
-   Export the refined sgRNA library (TSV file)
-   Export the total sgRNA or gene lost (Excel file)


```{r Export refined sgRNA library, include = FALSE}

# Export refined sgRNA library
value <- paste0(round(length(sgRNA_library_refined$sgRNA) / 1000, digits = 0), "K")
write_tsv(sgRNA_library_refined, 
          file = paste(params$output_dir, sgRNA_library_selected, "_", value, "_refined.tsv", sep = ""), col_names = FALSE)

write_tsv(list_discarded_sgrna, 
          file = paste(params$output_dir, sgRNA_library_selected, "_disposed_sgRNAs.tsv", sep = ""), col_names = TRUE)

write.xlsx(
  list("Genes Report" = total_sgRNA_or_gene,
       "Refining Process Tracker" = track_table),
  file = paste(params$output_dir, sgRNA_library_selected, "_full_report.xlsx", sep = "")
)

```

## Details of R session

```{r Show R session, echo = FALSE}

sessionInfo()

```



