---
title: "sgRNA Library Cleaning Pipeline"
author: "Author: Stefanus Bernard; small adaptation from Barbara & Ryan (2019)"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
    html_document:
      theme: cerulean
      toc: true
      toc_float: true
      number_sections: true
      self_contained: true
---

# Input data and Methods

## Input data
1. sgRNA library TSV file input for MAGeCK (Format: sgRNA, spacer, gene; without header)
2. Human genome GRCh38 indexed in bowtie
3. CCDS annotation file. Look for the current release here: https://ftp.ncbi.nlm.nih.gov/pub/CCDS/current_human/
4. HGNC (Hugo Gene Nomenclature Committee) annotation file. Look for the current release here: https://www.genenames.org/download/custom/

The current HGNC dataset were retrieved from https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt

## Methods

1.  Import required packages
2.  Define any filename
3.  Import sgRNA library and create dir for output (output_cleaning) and intermediate files (object_intermediate)
4.  Alignment of sgRNA to HG38 UCSC using Bowtie2 (2 mismatches)
5.  Post alignment data processing
6.  Find and drop multi-target sgRNAs
7.  Find and drop single mismatch sgRNAs
8.  Find and drop sgRNA align with double mismatches at the PAM distal nucleotides
9.  CCDS annotation of sgRNA library
10. HGNC annotation of sgRNA library
11. Clean sgRNA library
12. Discard any genes targeted by less than 3 sgRNAs
13. Calculate how many sgRNAs or genes lost
14. Export the refined sgRNA library TSV file and excel file report

# Pipeline

```{r install required packages, include = FALSE}

setwd(getwd())
source("sgRNA_library_refinement_pipeline_functions.R")

list_packages_bioconductor <- c("crisprBase",
                                "crisprBowtie",
                                "BSgenome.Hsapiens.UCSC.hg38",
                                "BSgenome.Hsapiens.NCBI.T2T.CHM13v2.0",
                                "GenomicRanges")

for (i in list_packages_bioconductor) {
  BiocManager::install(i)
  
}


list_packages <- c("hgnc",
                   "stringr",
                   "directlabels",
                   "tidyverse",
                   "knitr")

new_packages <- list_packages[!(list_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

```

```{r Import required library, include = FALSE}

library(crisprBase)
library(crisprBowtie)
library(Rbowtie)
library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Hsapiens.NCBI.T2T.CHM13v2.0)
library(GenomicRanges)
library(hgnc)
library(stringr)
library(directlabels)
library(tidyverse)
library(knitr)
library(xlsx)

```

## Detected file input

```{r User defined variable (User input/Hardcoded), echo = FALSE}

# main sgRNA library file location
file_dir_location <- "./public_crispr_library/"

# name of sgRNA library (without file format)
sgRNA_library_selected <- "broadgpp-brunello-library-contents"

# directory for NCBI CCDS data
ccds_filename_dir <- "CCDS_20250112.txt"

# name output for sgRNA library alignment (.CSV file)
aln_2_mismatch_filename <- paste(sgRNA_library_selected, "_aln.csv", sep ="")


# Bowtie alignment indexing
check_index_file = "hg38.1.ebwt"
bowtie_index_dir <- "bowtie_index/hg38/"
name_prefix_for_index <- "hg38"


paste("sgRNA library name:", sgRNA_library_selected)
paste("CCDS filename:", ccds_filename_dir)
paste("Alignmment with 2 mismatch filename (csv format):", aln_2_mismatch_filename)

```

```{r Create folder and import library cellecta, include = FALSE}
dir.create("output_cleaning")
dir.create("object_intermediate")
sgRNA_library <- read_tsv(paste(file_dir_location, sgRNA_library_selected, ".tsv", sep = ""), col_names = FALSE)
colnames(sgRNA_library) <- c("sgRNA", "spacer", "gene")

```

```{r Check for HG38 index files, include = FALSE}

list.files(path = bowtie_index_dir, pattern = "*.ebwt$")
file_test("-f", paste(bowtie_index_dir, check_index_file, sep="")) # returns TRUE

```

```{r Indexing HG38 if indexing have not been done previously, include = FALSE}
if (file_test("-f", paste(bowtie_index_dir, check_index_file, sep="")) == FALSE) {
  Rbowtie::bowtie_build(fastaFile,
                         outdir = bowtie_index_dir,
                         force = TRUE,
                         prefix = name_prefix_for_index)
} else {
  print(paste(check_index_file, "is in bowtie_index directory, indicating HG38 indexing by bowtie have been done previously"))
}
```

```{r Check whether alignment has been done, include = FALSE}

if (file_test("-f", paste("./object_intermediate/", aln_2_mismatch_filename, sep = "")) == FALSE) {
      data(SpCas9, package="crisprBase")
      crisprNuclease <- SpCas9
      
      spacers_sequence <- c(unique(sgRNA_library$spacer))
      
      bowtie_index <- file.path(bowtie_index_dir, name_prefix_for_index)
      
      alignment_guides <- runCrisprBowtie(spacers_sequence,
                                           mode = "spacer",
                                           bsgenome = BSgenome.Hsapiens.UCSC.hg38,
                                           crisprNuclease = crisprNuclease,
                                           n_mismatches = 2,
                                           force_spacer_length = TRUE,
                                           canonical = TRUE,
                                           bowtie_index = bowtie_index)
      
      alignment_data <- sgRNA_library %>%
        left_join(alignment_guides, by = join_by(spacer))
      
      write.csv(alignment_data, 
                file = paste("./object_intermediate/", aln_2_mismatch_filename, sep = ""), 
                row.names = FALSE)
  
} else {
  paste(aln_2_mismatch_filename, "file is in object_intermediate folder, this indicates the alignment with max 2 mismatches has been conducted previously")
}

```

```{r import csv file from the alignment, echo = FALSE, include = FALSE}

library_alignment <- read_csv(paste("./object_intermediate/", aln_2_mismatch_filename, sep = ""))

```

```{r, echo = FALSE, include = FALSE}

track_table <- data.frame(matrix(ncol = 4, nrow = 0))
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = "count beginning")
```

```{r Select original gene list and remove any alignment to non standard chromosome, include = FALSE}
gene_original <- c(unique(library_alignment$gene))

list_chromosome <- paste0("chr", c(1:22, "X", "Y"))
library_alignment <- library_alignment %>%
  filter(chr %in% list_chromosome | is.na(chr))

```

## Report of Multi-target sgRNAs

```{r Find multi-target sgRNAs, include = FALSE}

# detect any multi-target sgRNAs
result_multi_target_sgrna <- multi_target_sgrna(library_alignment)

# remove multi-target sgRNAs
library_alignment <- library_alignment %>%
  filter(!sgRNA %in% result_multi_target_sgrna$list_guides)

# genes lost due to multi-target sgRNAs
genes_lost_multi_target_sgrna <- setdiff(gene_original, unique(library_alignment$gene))

# differences between gene original and gene lost
gene_original <- setdiff(gene_original, genes_lost_multi_target_sgrna)

# number of gene lost due to multi-target sgRNAs
print(paste("Number of genes lost due to multi-target sgRNAs: ", length(genes_lost_multi_target_sgrna), sep = ""))

```

```{r, include = FALSE}
# update track table
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = paste(length(result_multi_target_sgrna$list_guides), " multi-target guides discarded"))
```

## Report of sgRNA aligned with single mismatch

```{r Find sgRNA aligned with single mismatch, include = FALSE}
# detect any single mismatch sgRNA
result_single_mismatch_sgrna <- single_mismatch_sgrna(library_alignment)

# remove single mismatch sgRNAs
library_alignment <- library_alignment %>%
  filter(!sgRNA %in% result_single_mismatch_sgrna$list_guides)

# genes lost due to single mismatch sgRNA
genes_lost_single_mismatch_sgrna <- setdiff(gene_original, unique(library_alignment$gene))

# differences between gene original and genes lost
gene_original <- setdiff(gene_original, genes_lost_single_mismatch_sgrna)

# number of genes lost due to single mismatch sgRNAs
print(paste("Number of genes lost due to single mismatch sgRNAs: ", length(genes_lost_single_mismatch_sgrna), sep = ""))
```

```{r, include = FALSE}
# update track table
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = paste(length(result_single_mismatch_sgrna$list_guides), " single mismatch sgRNA discarded"))
```

## Report of sgRNA aligned with double mismatches at the PAM distal

**sgRNAs aligned with double mismatch at PAM distal site may able to tolerate the off-target activity**

```{r Find sgRNAs with 2 mismatches alignments, include = FALSE}

# detect any pam-distal double mismatch sgRNA
result_pam_distal_double_mismatch <- pam_distal_double_mismatch(library_alignment)

# remove pam-distal double mismatch sgRNA
library_alignment <- result_pam_distal_double_mismatch$pam_distal_df %>%
  filter(!sgRNA %in% result_pam_distal_double_mismatch$list_guides)

# genes lost due to pam-distal double mismatch sgRNA
genes_lost_pam_distal_double_mismatch_sgrna <- setdiff(gene_original, unique(library_alignment$gene))

# differences between gene original and genes lost
gene_original <- setdiff(gene_original, genes_lost_pam_distal_double_mismatch_sgrna)

# number of genes lost due to single mismatch sgRNAs
print(paste("Number of genes lost due to PAM-distal double mismatch sgRNAs: ", length(genes_lost_pam_distal_double_mismatch_sgrna), sep = ""))

```

## Compiled report of multi-target, single mismatch, and PAM distal double mismatch sgRNAs

```{r, echo = FALSE}
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = paste(length(result_pam_distal_double_mismatch$list_guides), " PAM-distal double mismatch sgRNA discarded"))
```

```{r Separate control or intron control sgRNA with the rest of sgRNA, include = FALSE}

# detect any control/intron control sgRNA
terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron")
terms_for_filtering <- paste(terms, collapse = "|")

# separate control sgRNA from the rest of sgRNA
sgRNA_library_control <- library_alignment %>%
  filter(str_detect(gene, terms_for_filtering)) %>%
  select(sgRNA, spacer, gene) %>%
  distinct()

# separate sgRNA not targeting anywhere
sgRNA_not_target_anywhere <- library_alignment %>%
  filter(is.na(protospacer) & !str_detect(gene, terms_for_filtering)) %>%
  select(sgRNA, gene) %>%
  distinct() %>%
  count(gene) %>%
  dplyr::rename("sgRNA not target anywhere" = "n")

# filter gene not in control and drop sgRNAs not targeting anywhere
library_alignment <- library_alignment %>%
  filter(!gene %in% unique(sgRNA_library_control$gene)) %>%
  drop_na()

# detected control sgRNA
detected_control_sgrna <- unique(sgRNA_library_control$gene)

```

```{r, echo = FALSE}
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(library_alignment$sgRNA)),
                                 num_genes = length(unique(library_alignment$gene)),
                                 num_alignments = length(library_alignment$sgRNA),
                                 notes = paste(length(unique(sgRNA_not_target_anywhere$sgRNA)), " sgRNAs not target anywhere discarded"))
```

```{r Using NCBI CCDS data to find the overlap of sgRNA alignment with exon annotation, include = FALSE}

# import ccds data
ccds <- read_ccds_data(ccds_filename = ccds_filename_dir)
# transform gene annotation data
transformed_gene_annotation <- transform_gene_annotation_ccds(ccds)

# take ccds_exon data from transformed gene annotation
ccds_exon <- transformed_gene_annotation$ccds_exon

# take granges data from transformed gene annotation
gene_annot_granges <- transformed_gene_annotation$gene_annot_granges_df

# add alignment id for every sgRNA alignment
library_alignment <- library_alignment %>% 
  mutate(unique_aln_id = paste0("sgrna_aln_", row_number())) %>%
  relocate(unique_aln_id, .before = sgRNA)

# make granges from alignment data
guide_aln_granges <- make_granges_from_alignment_data(library_alignment)

# final result of gene_df is a dataframe containing the overlap of gene annotation and alignment
gene_df <- find_overlaps_gene_annotation_and_alignment(guide_aln_granges = guide_aln_granges,
                                                       gene_annot_granges = gene_annot_granges)

```

## Validation of gene symbol for the alignment and annotation data

**Alignment validation to check whether there is a symbol mismatch between the gene symbol and the exon annotation**

```{r Validation of gene symbol for the alignment and annotation data - Strictly pick 0 mismatch alignment, echo = FALSE}

gene_df %>%
  filter(mismatches == 0) %>%
  mutate(notes = case_when(str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)") == gene ~ "symbol match",
                           str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)") != gene ~ "symbol mismatch")) %>%
  select(gene, notes) %>%
  distinct() %>%
  count(notes) %>%
  dplyr::rename("how_many" = "n") %>%
  ggplot(aes(notes, how_many)) +
    geom_histogram(stat = "identity") +
    labs(
      title = paste("Validation of sgRNA alignment with exon and gene symbol annotation"),
      x = "",
      y = "Number of genes"
    ) +
    geom_text(aes(label = how_many), vjust=-0.5) +
    scale_x_discrete(labels=c("symbol match" = "gene symbol match with exon annotation",
                              "symbol mismatch" = "gene symbol not match \nwith exon annotation")) +
    theme_light()

```

**Correct the sgRNA according to targeted gene symbol**

```{r Correct the sgRNA according to targeted gene symbol, echo = FALSE}

# categorize sgRNA with gene symbol match and sgRNA with gene symbol mismatch
gene_df_check <- gene_df %>%
  filter(mismatches == 0) %>%
  mutate(notes = case_when(str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)") == gene ~ "symbol match",
                           str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)") != gene ~ "symbol mismatch"))

# check for duplicate spacer in gene_df_match and discard it (sgRNA targeting the same gene but different exon location)
gene_df_match <- gene_df_check %>%
  filter(notes == "symbol match") %>%
  mutate(duplicate_spacer_gene = duplicated(spacer, gene)) %>%
  arrange(sgrna) %>%
  filter(!duplicate_spacer_gene == TRUE)

# separate sgRNA with gene symbol mismatch dataframe, discard duplicated spacer
gene_df_mismatch <- gene_df_check %>%
  filter(notes == "symbol mismatch") %>%
  mutate(duplicate_spacer_gene = duplicated(spacer, gene)) %>%
  arrange(gene) %>%
  filter(!duplicate_spacer_gene == TRUE)

# arrange sequential numbers here
run_length <- rle(gene_df_mismatch$gene)$lengths
sequential_numbers <- sequence(run_length)

# correct sgRNA - gene name according to the alignment exon location and add new sgRNA number id alongside _SC (corrected gene symbol)
gene_df_mismatch <- gene_df_mismatch %>%
  mutate(new_sgrna_id = paste("sg", gene, "_", sequential_numbers, "_SC" , sep ="")) %>%
  select(-sgrna) %>%
  dplyr::rename("sgrna" = "new_sgrna_id") %>%
  relocate(sgrna, .before = spacer) %>%
  mutate(notes = "symbol corrected")

# find corrected genes that already exist in gene_df_match put it into a new dataframe called as corrected_exist_gene
list_corrected_exist_gene <- intersect(gene_df_match$gene, gene_df_mismatch$gene)
corrected_exist_gene <- gene_df_match %>%
  filter(gene %in% list_corrected_exist_gene)

# filter out problematic gene (gene that were targeted by sgRNA in Cellecta library, but also targeted by another sgRNA with corrected symbol)
gene_df_match <- gene_df_match %>%
  filter(!gene %in% list_corrected_exist_gene)

# join gene_df_mismatch with gene_df_match to find the duplicated sgRNA
gene_df_mismatch <- gene_df_mismatch %>%
  full_join(corrected_exist_gene) %>%
  arrange(gene, desc(notes)) %>%
  mutate(duplicate_spacer_gene = duplicated(spacer, gene))

# full_join the gene_df_mismatch with gene_df_match to become a gene_df_corrected
gene_df_corrected <- gene_df_match %>%
  full_join(gene_df_mismatch)

# Histogram of corrected gene symbol
gene_df_corrected %>%
  filter(mismatches == 0) %>%
  mutate(notes = case_when(str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)") == gene ~ "symbol match",
                           str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)") != gene ~ "symbol mismatch")) %>%
  select(gene, notes) %>%
  distinct() %>%
  count(notes) %>%
  dplyr::rename("how_many" = "n") %>%
  ggplot(aes(notes, how_many)) +
    geom_histogram(stat = "identity") +
    labs(
      title = paste("Validation of sgRNA alignment with exon and gene symbol annotation"),
      x = "",
      y = "Number of genes"
    ) +
    geom_text(aes(label = how_many), vjust=-0.5) +
    scale_x_discrete(labels=c("symbol match" = "gene symbol match with exon annotation",
                              "symbol mismatch" = "gene symbol not match \nwith exon annotation")) +
    theme_light()

```

```{r Find genes lost after annotation, include = FALSE}

# gene_track_annotation : a dataframe of the old and approved gene symbol
gene_track_annotation <- gene_df_check %>%
  filter(notes == "symbol mismatch") %>%
  select(sgrna, gene, notes) %>%
  mutate(old_gene = str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)")) %>%
  relocate(old_gene, .before = gene) %>%
  select(-sgrna) %>%
  distinct()

prev_gene <- c(gene_track_annotation$old_gene)
align_gene <- c(gene_track_annotation$gene)

gene_original_corrected <- sapply(gene_original, function(gene_query){
  match_index <- match(gene_query, prev_gene)
  if (!is.na(match_index)) {
   align_gene[match_index]
  } else {
    gene_query
  }
})

gene_original_corrected <- as.character(gene_original_corrected)

paste("Corrected gene name: ")
setdiff(gene_original_corrected, gene_original)

exclude_control <- c(unique(gene_df_corrected$gene), detected_control_sgrna)

genes_lost_after_annotation <- setdiff(gene_original_corrected, exclude_control)
gene_original_corrected <- setdiff(gene_original_corrected, genes_lost_after_annotation)

```

[**Genes lost after annotation**]{style="color:red"}

**Genes lost at this stage means that its sgRNA was not aligned to anywhere in the genome**

```{r Genes lost after gene annotation, echo = FALSE}

print(paste("number of genes lost following the annotation: ", length(genes_lost_after_annotation), sep = ""))

```

```{r, include = FALSE}
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(gene_df_corrected$sgrna)),
                                 num_genes = length(unique(gene_df_corrected$gene)),
                                 num_alignments = length(gene_df_corrected$sgrna),
                                 notes = paste(length(unique(gene_df_corrected$sgrna)), " sgRNA targeting exon only"))
```

## Filter only sgRNAs targeting exons of protein-coding genes

```{r Find sgRNAs failed to target exon, include = FALSE}

list_sgRNA_not_target_exon <- setdiff(unique(library_alignment$sgRNA), unique(gene_df_corrected$sgrna))

sgRNA_not_target_exon <- library_alignment %>%
  select(sgRNA, gene) %>%
  filter(sgRNA %in% list_sgRNA_not_target_exon) %>%
  distinct() %>%
  count(gene) %>%
  filter(!gene %in% gene_track_annotation$old_gene) %>%
  dplyr::rename("sgRNA not target exon" = "n")

```

```{r Add HGNC annotation of sgRNA library, include = FALSE}
hgnc <- read_tsv(url("https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt"))

hgnc_selected <- hgnc %>%
  select(symbol, name, locus_group, locus_type) %>%
  dplyr::rename("gene" = "symbol")

sgRNA_library_gene_df <- gene_df_corrected %>%
  left_join(hgnc_selected, by = "gene")

paste("count sgRNAs: ", length(unique(sgRNA_library_gene_df$sgrna)))
paste("alignments left after data joining with hgnc:", length(sgRNA_library_gene_df$sgrna))

```

```{r Filter sgRNAs targeting protein-coding genes only, include = FALSE}

sgRNA_library_gene_df_annot <- sgRNA_library_gene_df %>%
  filter(locus_group == "protein-coding gene")

paste("count sgRNAs: ", length(unique(sgRNA_library_gene_df_annot$sgrna)))
paste("alignments left after data selection of protein-coding gene only:", length(sgRNA_library_gene_df_annot$sgrna))

```

```{r Find genes failed in the annotation, include = FALSE}

exclude_control_2 = c(unique(sgRNA_library_gene_df_annot$gene), detected_control_sgrna)
genes_not_pcg <- setdiff(sgRNA_library_gene_df$gene, exclude_control_2)

genes_not_pcg_df <- sgRNA_library_gene_df %>%#
  filter(gene %in% genes_not_pcg)
  
```

[**sgRNAs not targeting protein coding genes**]{style="color:red"}

**sgRNAs lost at this stage means that the sgRNA is not targeting a protein-coding genes**

sgRNAs might target:

-   Readthrough gene

-   Complex locus constituent

-   Pseudogene

```{r Selected only protein-coding genes, echo = FALSE}

print(paste("number of non protein-coding genes: ", length(genes_not_pcg), sep = ""))
# print(genes_not_pcg)

```

```{r, include = FALSE}
track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(sgRNA_library_gene_df_annot$sgrna)),
                                 num_genes = length(unique(sgRNA_library_gene_df_annot$gene)),
                                 num_alignments = length(sgRNA_library_gene_df_annot$sgrna),
                                 notes = paste(length(unique(sgRNA_library_gene_df_annot$sgrna)), " sgRNA successfully annotated"))

```

```{r Obtain refined sgRNA library, echo = FALSE, include = FALSE}

sgRNA_library_cleaned <- sgRNA_library_gene_df_annot %>%
  select(sgrna, spacer, gene) %>%
  distinct() %>%
  dplyr::rename("sgRNA" = "sgrna")

count_guides_per_gene <- sgRNA_library_cleaned %>%
  count(gene) %>%
  count(n) %>%
  mutate(`total number of guides` = n * nn) %>%
  dplyr::rename("number of guides per gene" = "n") %>%
  dplyr::rename("how many genes" = "nn")

print(paste("total number of sgRNA: ", sum(count_guides_per_gene$`total number of guides`)))

num_original_genes <- length(unique(sgRNA_library$gene))
num_genes_after_cleaning <- sum(count_guides_per_gene$`how many genes`)

paste("number of original target genes in the library: ", length(unique(sgRNA_library$gene)))
paste("number of genes after cleaning: ", sum(count_guides_per_gene$`how many genes`))
paste("number of genes lost: ", num_original_genes-num_genes_after_cleaning)

```

```{r Discard any genes targeted by < 3 sgRNAs, echo = FALSE, include = FALSE}

guides_below_3 <- sgRNA_library_cleaned %>%
  count(gene) %>%
  filter(n < 3)

sgRNA_library_cleaned <- sgRNA_library_cleaned %>%
  filter(!gene %in% guides_below_3$gene)

count_guides_per_gene <- sgRNA_library_cleaned %>%
  count(gene) %>%
  count(n) %>%
  mutate(`total number of guides` = n * nn) %>%
  dplyr::rename("number of guides per gene" = "n") %>%
  dplyr::rename("how many genes" = "nn")

num_original_genes <- length(unique(sgRNA_library$gene))
num_genes_after_cleaning_3_guides <- length(unique(sgRNA_library_cleaned$gene))

paste("total number of sgRNA: ", sum(count_guides_per_gene$`total number of guides`))
paste("number of original target genes in the library: ", length(unique(sgRNA_library$gene)))
paste("number of genes after exclude gene targeted by guides < 3: ", length(unique(sgRNA_library_cleaned$gene)))
paste("number of genes lost: ", num_original_genes-num_genes_after_cleaning_3_guides)

```

## Final report of refined sgRNA library

```{r, echo = FALSE}

track_table <- function_df_count(track_table,
                                 count_sgrna = length(unique(sgRNA_library_cleaned$sgRNA)),
                                 num_genes = length(unique(sgRNA_library_cleaned$gene)),
                                 num_alignments = "-",
                                 notes = paste(length(unique(guides_below_3$gene)), " genes discarded (< 3 sgRNAs)"))

colnames(track_table) <- c("count_sgrna", "num_genes", "num_alignments", "notes")
kable(track_table)
```

```{r Combine the control sgRNA back with the original dataframe, include = FALSE}	

# Add control sgRNA (sgRNA_library_control)
sgRNA_library_refined <- rbind(sgRNA_library_control, sgRNA_library_cleaned)

```

```{r Calculate total sgRNAs and genes lost, include = FALSE}

# filter(str_detect(gene, terms_for_filtering))

total_sgRNA_or_gene_lost <- data.frame("gene" = sgRNA_library$gene) %>%
  count(gene) %>%
  dplyr::rename("sgRNA number" = "n") %>%
  left_join(result_multi_target_sgrna$dropped_genes_num, by = "gene") %>%
  mutate(`multi-target sgRNA` = replace_na(`multi-target sgRNA`, 0)) %>%
  left_join(result_single_mismatch_sgrna$dropped_genes_num, by = "gene") %>%
  mutate(`single mismatch sgRNA` = replace_na(`single mismatch sgRNA`, 0)) %>%
  left_join(result_pam_distal_double_mismatch$dropped_genes_num, by = "gene") %>%
  mutate(`PAM distal double mismatch` = replace_na(`PAM distal double mismatch`, 0)) %>%
  left_join(sgRNA_not_target_anywhere, by = "gene") %>%
  mutate(`sgRNA not target anywhere` = replace_na(`sgRNA not target anywhere`, 0)) %>%
  left_join(sgRNA_not_target_exon, by = "gene") %>%
  mutate(`sgRNA not target exon` = replace_na(`sgRNA not target exon`, 0)) %>%
  mutate(
    `counted total sgRNA` = `sgRNA number` - (`multi-target sgRNA` + `single mismatch sgRNA` + `PAM distal double mismatch` + `sgRNA not target anywhere` + `sgRNA not target exon`)) %>%
  mutate(`symbol change` = case_when(gene %in% gene_track_annotation$old_gene ~
                                       gene_track_annotation$gene[match(gene, gene_track_annotation$old_gene)],
                                        TRUE ~ "no")) %>%
  mutate(`protein-coding gene` = case_when(str_detect(gene, terms_for_filtering) ~ "no",
                                           gene %in% genes_not_pcg ~ "no",
                                           TRUE ~ "yes"))

crosscheck_count <- sgRNA_library_refined %>%
  count(gene) %>%
  dplyr::rename("actual total sgRNA" = "n")

total_sgRNA_or_gene <- total_sgRNA_or_gene_lost %>%
  left_join(crosscheck_count, by = join_by(gene)) %>%
  mutate(`actual total sgRNA` = replace_na(`actual total sgRNA`, 0)) %>%
  relocate(`actual total sgRNA`, .after = `counted total sgRNA`) %>%
  mutate(`number match` = mapply(identical,                 
                                 `counted total sgRNA`, 
                                 `actual total sgRNA`)) %>%
  relocate(`number match`, .after = `actual total sgRNA`) %>%
  mutate(`quality check` = case_when((`symbol change` != "no") ~ "not counted (symbol change)",
                                     (`actual total sgRNA` > 2 & `protein-coding gene` == "yes") ~ "pass",
                                     TRUE ~ "fail")) %>%
  mutate(`quality check` = ifelse(gene %in% unique(sgRNA_library_control$gene), "pass", `quality check`)) %>%
  arrange(factor(gene, levels = c(unique(sgRNA_library_control$gene), setdiff(unique(gene), unique(sgRNA_library_control$gene)))))

```

## Export the refined library and sgRNA cleaning report
-   Export cleaned sgRNA library (TSV file)
-   Export total sgRNA or gene lost (Excel file)

```{r Snippets of exported file, echo = FALSE}

kable(head(sgRNA_library_refined, 25))
kable(head(total_sgRNA_or_gene, 25))

```

```{r Export cleaned sgRNA library, include = FALSE}

# Export refined sgRNA library
value <- paste0(round(length(sgRNA_library_refined$sgRNA) / 1000, digits = 0), "K")
write_tsv(sgRNA_library_refined, 
          file = paste("./output_cleaning/", sgRNA_library_selected, "_", value, "_refined.tsv", sep = ""), col_names = FALSE)

write.xlsx(total_sgRNA_or_gene, file = paste("./output_cleaning/", sgRNA_library_selected, "_full_report.xlsx", sep = ""), row.names = FALSE, sheetName = "Genes Report", append = TRUE)
write.xlsx(track_table, file = paste("./output_cleaning/", sgRNA_library_selected, "_full_report.xlsx", sep = ""), row.names = FALSE, sheetName = "Refining Process Tracker", append = TRUE)


```

