---
title: "sgRNA Library Pre-Cleaning Script"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

# Important Notes

## Hypothesis

CRISPR screen library often has discrepancy of gene name with the updated HGNC dataset

## Objective

Correcting the gene name of CRISPR screen library

## Materials and Methods

### Materials

1.  CRISPR screen library
2.  HGNC dataset

### Methods

1.  Import HGNC dataset
2.  Import sgRNA CRISPR library
3.  Correction of sgRNA library

# Experiment

## Import required packages

```{r}
library(hgnc)
library(tidyverse)
library(naniar)
```

## Import sgRNA CRISPR library and define control sgRNA (user input)

```{r}
sgRNA_library_selected <- 'Cellecta'
sgRNA_library <- read_csv(paste('./raw_data/', sgRNA_library_selected, '.csv', sep = ''))

# should be user input later
control_sgrna <- c('Neg_Control', 'INTRON_CTRL')
```

length(unique(sgRNA_library_ori\$gene)) *19,013 genes originally detected*

## Import HGNC dataset

```{r}
hgnc <- import_hgnc_dataset()

hgnc_selected <- hgnc %>%
  select(symbol, prev_symbol, alias_symbol, location_sortable)

hgnc_selected

```

## Check how many missing genes

Crosscheck_genes is only the 2x2 table to show TRUE or FALSE

```{r}

`%notin%` <- Negate(`%in%`)

crosscheck_genes <- sgRNA_library %>%
  filter(gene %notin% control_sgrna)

crosscheck_genes <- data.frame(genes_in_hgnc = unique(crosscheck_genes$gene) %in% unique(hgnc_selected$symbol))
crosscheck_genes <- crosscheck_genes %>%
  count(genes_in_hgnc) %>%
  dplyr::rename('how_many' = 'n')

ggplot(crosscheck_genes, aes(genes_in_hgnc, how_many)) +
  geom_histogram(stat = 'identity') +
  labs(
    title = paste('Number of discrepancy between', sgRNA_library_selected, 'sgRNA library and gene annotation from HGNC'),
    x = '',
    y = 'Number of genes'
  ) +
  geom_text(aes(label = how_many), vjust=-0.5) +
  scale_x_discrete(labels=c('TRUE' = 'match with HGNC data',
                            'FALSE' = 'missing gene \n(unmatched with HGNC data)')) +
  theme_light()

```

## Fetch missing genes

```{r}

missing_genes <- sgRNA_library %>%
  filter(gene %notin% hgnc_selected$symbol) %>%
  filter(gene %notin% control_sgrna)
  
```

## Separate original library/crosschecked genes with missing genes

```{r}

sgRNA_library <- sgRNA_library %>%
  filter(gene %notin% missing_genes$gene)

paste(length(unique(sgRNA_library$gene)), 'genes left after separation of sgRNA library data from missing genes')

```

## Find the missing genes by crosschecking with HGNC data

```{r}

find_missing_genes <- data.frame(missing_gene_name = unique(missing_genes$gene))

find_missing_genes <- find_missing_genes %>%
  mutate(hgnc_current_gene_name = sapply(missing_gene_name, 
                                         function(gene) {
                                           match_index <- which(sapply(hgnc_selected$prev_symbol, function(symbol_list) gene %in% symbol_list))
                                           if (length(match_index) == 0) {
                                             match_index <- which(sapply(hgnc_selected$alias_symbol,
                                                                         function(symbol_list) gene %in% symbol_list))}
                                           if (length(match_index) == 1) {
                                             hgnc_selected$symbol[match_index][1]
                                           }
                                           else if (length(match_index) > 1){
                                               hgnc_selected$symbol[match_index]
                                           }
                                           else {
                                             NA
                                           }
                                           }))


# special genes is those gene with more than one approved symbol
special_genes <- find_missing_genes %>%
  filter(sapply(hgnc_current_gene_name, length) > 1)

# separate find_missing_genes from the gene with more than one approved symbol
find_missing_genes <- find_missing_genes %>%
  mutate(hgnc_current_gene_name = sapply(hgnc_current_gene_name, function(x) {
    if (length(x) > 1) {
      return(NA)
    } else {
      return(x)
    }
  }))
  
paste('how many genes missing:' , length(unique(find_missing_genes$missing_gene_name)))
# how many genes missing: 1463

paste('how many genes successfully crossmatched with HGNC data:', length(unique(find_missing_genes$hgnc_current_gene_name)))
# how many genes successfully crossmatched with HGNC data: 1343


# create a dataframe to check how many genes successfully match with HGNC data and those that are still missing
find_missing_gene_df <- data.frame(failed_crosscheck = is.na(find_missing_genes$hgnc_current_gene_name))
find_missing_gene_df <- find_missing_gene_df %>%
  count(failed_crosscheck) %>%
  dplyr::rename('how_many_genes' = 'n')


# plot how many data successfully crossmatched (excluding the special genes and missing genes after crossmatching)
ggplot(find_missing_gene_df, aes(failed_crosscheck, how_many_genes)) +
  geom_histogram(stat = 'identity') +
  labs(
    title = paste('Comparison between missing genes and successfully annotated gene from HGNC'),
    x = '',
    y = 'Number of genes'
  ) +
  geom_text(aes(label = how_many_genes), vjust=-0.5) +
  scale_x_discrete(labels=c('TRUE' = 'missing gene \n(unmatched with HGNC data)',
                            'FALSE' = 'gene match with HGNC data')) +
  theme_light()

```

## Detect any duplicated genes

```{r}

duplicated_genes <- find_missing_genes %>%
  count(hgnc_current_gene_name) %>%
  filter(n > 1) %>%
  dplyr::rename('num_genes' = 'n')

duplicated_genes_df <- find_missing_genes %>%
  filter(hgnc_current_gene_name %in% duplicated_genes$hgnc_current_gene_name,
         !is.na(hgnc_current_gene_name))

```

## Separate genes that failed during crosschecking and those successfully crosschecked

```{r}

# separate genes that failed crosschecking with HGNC dataset
genes_unknown <- find_missing_genes %>%
  filter(is.na(hgnc_current_gene_name))

# rename the missing_gene_name column to 'gene' to match the missing_genes dataframe
find_missing_genes <- find_missing_genes %>%
  dplyr::rename('gene' = 'missing_gene_name')

```

## Analysis of unknown genes

```{r}

genes_unknown <- genes_unknown %>%
  mutate(notes = case_when(missing_gene_name %in% special_genes$missing_gene_name ~ 'ambiguous annotation',
                str_detect(missing_gene_name, 'LOC') ~ 'uncharacterized gene',
                str_detect(missing_gene_name, '-') ~ 'possibly targeting two genes',
                TRUE ~ 'unknown'))

count_genes_unknown <- genes_unknown %>%
  count(notes) %>%
  dplyr::rename('number_of_genes' = 'n')

ggplot(count_genes_unknown, aes(notes, number_of_genes)) +
  geom_histogram(stat = 'identity') +
  labs(
    title = paste('Further Analysis of Missing Genes'),
    x = '',
    y = 'Number of genes'
  ) +
  geom_text(aes(label = number_of_genes), vjust=-0.5) +
  scale_x_discrete(labels=c('TRUE' = 'missing gene \n(unmatched with HGNC data)',
                            'FALSE' = 'gene match with HGNC data')) +
  theme_light()

```

## Map separated data and replace sgRNA gene name with the recent HGNC version gene name

```{r}

# left join the find_missing_genes with missing_genes data
missing_genes <- missing_genes %>%
  left_join(find_missing_genes, by = join_by('gene')) %>%
  mutate(new_sgrna = str_replace(sgRNA, "(?<=sg)([A-Za-z0-9-]+)", hgnc_current_gene_name))

paste('number of genes in missing_genes:', length(unique(missing_genes$gene)))
# number of genes in missing_genes: 1463

final_corrected_missing_genes <- missing_genes %>%
  mutate(hgnc_current_gene_name = dplyr::coalesce(hgnc_current_gene_name, gene),
         new_sgrna = dplyr::coalesce(new_sgrna, sgRNA)) %>%
  select(new_sgrna, hgnc_current_gene_name, spacer) %>%
  dplyr::rename('sgRNA' = 'new_sgrna',
                'gene' = 'hgnc_current_gene_name')

paste('number of genes in final_corrected_missing_genes:', length(unique(final_corrected_missing_genes$gene)))
# number of genes in final_corrected_missing_genes: 1451


# sgrna <- 'sgAARS1_1'
# paste(str_extract(sgrna, "(?<=sg)([A-Za-z0-9-]+)"), sep = '_')
  
```

## Combine the dataframe of crosschecked genes with original dataframe

```{r}

length(unique(final_corrected_missing_genes$gene))
# 1451 genes
length(unique(sgRNA_library$gene))
# 17550 genes

sgRNA_library_final <- sgRNA_library %>%
  full_join(final_corrected_missing_genes)

length(unique(sgRNA_library_final$gene))
# 18933 genes

# 19013 - 18929 = 80 genes lost

# check how many count genes in final dataframe
count_gene_final <- sgRNA_library_final %>%
  count(gene) %>%
  filter(n < 80) %>%
  dplyr::rename('sgrna_num' = 'n')

```

## Plot comparison before and after the cleaning

```{r}
# how many NA in gene column
sum(is.na(sgRNA_library_final$gene))

# how many NA in sgRNA column
sum(is.na(sgRNA_library_final$sgRNA))

# any missing genes left ? cross-compare with HGNC dataset
length(unique(sgRNA_library_final$gene))
# 18933 genes

# create dataframe for final_check_count; compare with crosscheck_genes

final_check_count <- data.frame(genes_name = unique(sgRNA_library_final$gene),
                                genes_in_hgnc = unique(sgRNA_library_final$gene) %in% unique(hgnc$symbol))

# exclude control sgRNA

final_check_count <- final_check_count %>%
  filter(genes_name %notin% control_sgrna) %>%
  count(genes_in_hgnc) %>%
  dplyr::rename('how_many' = 'n')

# combine crosscheck_genes and final_check_count

crosscheck_genes$genes_in_hgnc = ifelse(crosscheck_genes$genes_in_hgnc, 'match hgnc', 'missing')
crosscheck_genes$group = 'before correction'
  
final_check_count$genes_in_hgnc = ifelse(final_check_count$genes_in_hgnc, 'match hgnc', 'missing')
final_check_count$group = 'after correction'

correction_table <- crosscheck_genes %>%
  full_join(final_check_count)

correction_table$group = factor(correction_table$group, levels = c('before correction', 'after correction'), ordered = TRUE)

ggplot(correction_table, aes(x = group, y = how_many, fill = genes_in_hgnc)) +
  geom_col(color = 'white') +
  labs(
    title = paste('Before-After Correction of Gene Symbols'),
    x = '',
    y = 'Number of genes') +
  geom_text(aes(label = how_many), vjust=-0.5) +
  theme_light() +
  scale_fill_manual(values=c('lightgreen', 'red'))


```

## Separate duplicated sgRNA and genes from final library

```{r}

# separate sgRNA library duplicated, and sort it based on the gene column
sgRNA_library_duplicated <- sgRNA_library_final %>%
  filter(gene %in% count_gene_final$gene) %>%
  arrange(gene)

# count the length of each gene and create sequential number
run_length <- rle(sgRNA_library_duplicated$gene)$lengths
sequential_numbers <- sequence(run_length)

sgRNA_library_duplicated <- sgRNA_library_duplicated %>%
  mutate(new_sgrna = paste('sg', gene, '_', sequential_numbers, sep = '')) %>%
  select(-sgRNA) %>%
  dplyr::rename('sgRNA' = 'new_sgrna') %>%
  dplyr::relocate(sgRNA, .before = gene)

```

## Combine the processed duplicated sgRNA back into the sgRNA_library_final

```{r}

sgRNA_library_min_duplicate <- sgRNA_library_final %>%
  filter(gene %notin% sgRNA_library_duplicated$gene)

sgRNA_library_export <- sgRNA_library_min_duplicate %>%
  full_join(sgRNA_library_duplicated)

```

## Show any genes targeted more than 8 sgRNAs

```{r}

sgrna_gene_duplicated <- sgRNA_library_export %>%
  count(gene) %>%
  dplyr::rename('sgrna_num' = 'n') %>%
  filter(sgrna_num > 8 & gene %notin% control_sgrna) %>%
  arrange(sgrna_num)

ggplot(sgrna_gene_duplicated, aes(gene, sgrna_num)) +
  geom_histogram(stat = 'identity') +
  labs(
    title = 'Genes targeted more than 8 sgRNAs',
    x = '',
    y = 'Number of sgRNA'
  ) +
  geom_text(aes(label = sgrna_num), vjust = -0.5) +
  theme(axis.text.x = element_text(vjust = 0.5, hjust = 0.5, angle = 45))

```

## Check for the final time for any duplicated sgRNA and gene in sgRNA_library_export

```{r}

paste('total unique sgRNA: ', length(unique(sgRNA_library_export$sgRNA)))
# total unique sgRNA:  150076

paste('total unique gene: ', length(unique(sgRNA_library_export$gene)))
# total unique gene: 18933

```

## Export corrected sgRNA library

```{r}

write_csv(sgRNA_library_export, file = paste('./object_intermediate/', sgRNA_library_selected, '_precleaning.csv', sep = ''))

```

# Result

# Conclusion
