---
title: "Reformatting Publicly Available sgRNA Library for MAGeCK input"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

## Input data
1. Publicly available CRISPR knockout sgRNA library downloaded from https://www.addgene.org/crispr/libraries/

## Methods

1.  Import required packages
2.  Import HGNC dataset
3.  Specify the directory of sgRNA library file
4.  Function to detect Gene symbol and Sequence column
5.  Function to re-format the sgRNA library according to MAGeCK input
6.  Function to export the table according to MAGeCK input TSV format
7.  Iterate to all files in public_crispr_library directory

```{r setup, include=FALSE}

library(tidyverse)
library(readxl)

```

````{r Import HGNC dataset, include=FALSE}

hgnc <- read_tsv(url('https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt'))
hgnc_gene <- hgnc$symbol

```

```{r Specify the directory where you put the sgRNA library file}

crispr_library_dir = './public_crispr_library/'

```

## Detected file input
**This code will detect any raw sgRNA library file in public_crispr_library directory**

```{r Function to detect Gene symbol and Sequence column, include=FALSE} 

function_detect_gene_and_sequence_col <- function(library_table, hgnc_symbol) {
  # detect sgRNA 20 nucleotide sequence based on the average number length of characters in each column
  average_char_length <- sapply(library_table, function(col) {
    mean(nchar(as.character(col)), na.rm = TRUE)
    })
  sequence_column <- colnames(library_table)[which(average_char_length == 20)]

  # detect gene symbol column based on the sampling of 30 rows of dataset and crossmatching with gene symbol column
  library_table_sample <- head(library_table, n = 30)
  matches <- sapply(library_table_sample, function(col) {
    any(col %in% hgnc_symbol)
  })
  gene_column <- colnames(library_table_sample)[which(matches == TRUE)]

  fetch_this_column <- c(gene_column, sequence_column)
  return(fetch_this_column)
}

```

```{r Function to re-format the sgRNA library according to MAGeCK input, include=FALSE}

`%notin%` <- Negate(`%in%`)

function_reformat_sgrna_library <- function(imported_library_table, detected_sgRNA_gene_column) {
  library_table <- imported_library_table %>%
    select(all_of(detected_sgRNA_gene_column)) %>%
    dplyr::rename(gene = !!sym(detected_sgRNA_gene_column[1]),
                  sequence = !!sym(detected_sgRNA_gene_column[2]))

  run_length <- rle(library_table$gene)$lengths
  sequential_number <- sequence(run_length)

  processed_library_table <- library_table %>%
    mutate(sgRNA = paste('sg', gene, '_', sequential_number, sep = '')) %>%
    dplyr::relocate(sgRNA, .before = gene) %>%
    dplyr::relocate(gene, .after = sequence)

  return(processed_library_table)
}

```

```{r Function to export the table according to MAGeCK input TSV format}

function_export_table <- function(library_table, filename) {
  write_tsv(library_table, filename, col_names = FALSE)
}

```


```{r Iterate to all files in public_crispr_library directory, include=FALSE}

for (file in dir(path = crispr_library_dir)) {
  if (grepl("\\.txt$", file)) {
    print(paste("Processing:", file))
    exported_filename <- sub("\\..*$", "", file)
    sgRNA_library <- read_tsv(paste(crispr_library_dir, file, sep = ''))
    sgRNA_and_sequence_column <- function_detect_gene_and_sequence_col(sgRNA_library, hgnc_gene)
    MAGeCK_format_sgRNA_library <- function_reformat_sgrna_library(sgRNA_library, sgRNA_and_sequence_column)
    function_export_table(MAGeCK_format_sgRNA_library, filename = paste(crispr_library_dir, exported_filename, '.tsv', sep = ''))

  } else if (grepl("\\.csv$", file)) {
    print(paste("Processing:", file))
    exported_filename <- sub("\\..*$", "", file)
    sgRNA_library <- read_csv(paste(crispr_library_dir, file, sep = ''))
    sgRNA_and_sequence_column <- function_detect_gene_and_sequence_col(sgRNA_library, hgnc_gene)
    MAGeCK_format_sgRNA_library <- function_reformat_sgrna_library(sgRNA_library, sgRNA_and_sequence_column)
    function_export_table(MAGeCK_format_sgRNA_library, filename = paste(crispr_library_dir, exported_filename, '.tsv', sep = ''))
  
  } else if (grepl("\\.xlsx$", file)) {
    print(paste("Processing:", file))
    exported_filename <- sub("\\..*$", "", file)
    sgRNA_library <- read_excel(paste(crispr_library_dir, file, sep = ''))
    sgRNA_and_sequence_column <- function_detect_gene_and_sequence_col(sgRNA_library, hgnc_gene)
    MAGeCK_format_sgRNA_library <- function_reformat_sgrna_library(sgRNA_library, sgRNA_and_sequence_column)
    function_export_table(MAGeCK_format_sgRNA_library, filename = paste(crispr_library_dir, exported_filename, '.tsv', sep = ''))
  }

}

```