---
title: "A script to convert GFF to CCDS format"
author: "Stefanus Bernard"
date: "`r Sys.Date()`"
output: html_document
---

# Import library

```{r setup, include=FALSE}

library(tidyverse)
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(ape)

```

# This is how to read CCDS files and convert it to a format where we could see the exon location in each CDS

```{r}

ccds <- read_delim("CCDS.20221027.txt", delim = '\t')

ccds$cds_from <- as.integer(ccds$cds_from)
ccds$cds_to <- as.integer(ccds$cds_to)

list_chr <- paste0('chr', c(1:22, 'X', 'Y'))

ccds <- ccds %>%
    dplyr::rename('chr' = '#chromosome') %>%
    mutate(chr = str_c("chr", chr)) %>%
    filter(ccds_status %in% c("Public", "Reviewed, update pending", "Under review, update")) %>%
    filter(chr %in% list_chr, !is.na(cds_from), !is.na(cds_to))

genome_info <- GenomeInfoDb::Seqinfo(genome = "hg38")[list_chr]

ccds_exon <- ccds %>%
    mutate(cds_interval = str_replace_all(cds_locations, "[\\[\\]]", "") %>% str_split("\\s*,\\s*")) %>%
    tidyr::unnest(cds_interval) %>%
    group_by(gene, gene_id, cds_locations) %>%
    mutate(exon_code = ifelse(cds_strand=="+", 1:dplyr::n(), dplyr::n():1)) %>%
    ungroup() %>%
    dplyr::mutate(cds_start = str_extract(cds_interval, "^[0-9]+") %>% as.integer,
                    cds_end = str_extract(cds_interval, "[0-9]+$") %>% as.integer) %>%
    dplyr::select(gene, gene_id, chr, start=cds_start, end=cds_end, strand=cds_strand,
                    gene_start=cds_from, gene_end=cds_to, exon_code)



```

# NCBI T2T-CHM13 GFF file convert to a NCBI CCDS like format

filter data in command line using zgrep -P "\tmRNA\t" GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz | grep "tag=RefSeq Select" > T2T-CHM13v2_canonical_mrna.gff3

Canonical transcript were defined by obtaining the tag RefSeq Select in the original GFF file (RefSeq Select is the representative transcript for every protein-coding genes) 

```{r}

## 1. Read canonical transcripts directly from gzipped GFF
canonical <- fread(
  cmd = "zgrep -P '\tmRNA\t' GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz | grep 'tag=RefSeq Select'",
  sep = "\t", header = FALSE,
  col.names = c("seqid","source","type","start","end","score","strand","phase","attr")
)[, .(
  transcript_id = sub(".*ID=([^;]+);.*", "\\1", attr),
  gene_id       = sub(".*GeneID:([0-9]+).*", "\\1", attr),
  gene_symbol   = sub(".*gene=([^;]+).*", "\\1", attr),
  seqid, strand
)]

## 2. Assembly report for accession ↔ chr mapping
assembly_report <- read_tsv("T2T-CHM13v2_assembly_report.txt")
mapping <- assembly_report %>%
  select(chr = `Sequence-Name`, RefSeq_Accession = `RefSeq-Accn`)

## 3. Read CDS features from the same GFF
cds <- fread(
  cmd = "zgrep -P '\tCDS\t' GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz",
  sep = "\t", header = FALSE,
  col.names = c("seqid","source","type","start","end","score","strand","phase","attr")
)[, .(
  transcript_id = sub(".*Parent=([^;]+).*", "\\1", attr),
  seqid, start, end, strand
)]

## 4. Join CDS with canonical transcripts
cds_canonical <- merge(
  cds[, .(transcript_id, seqid, start, end, strand)],        # only keep unique columns from CDS
  canonical[, .(transcript_id, gene_id, gene_symbol)],       # drop seqid/strand here
  by = "transcript_id",
  allow.cartesian = TRUE
)

## 5. Collapse CDS ranges per gene
cds_gene <- cds_canonical[, .(
  CDS_ranges = list(paste0(start, "-", end)),  # proper list of ranges
  CDS_start  = min(start),
  CDS_end    = max(end)
), by = .(gene_id, gene_symbol, seqid, strand)]

## 6. Map accessions to chr names and tidy columns
cds_gene <- merge(cds_gene, mapping, by.x="seqid", by.y="RefSeq_Accession", all.x=TRUE)

cds_gene <- cds_gene %>%
  dplyr::rename(
    "nc_accession" = "seqid",
    "gene" = "gene_symbol",
    "cds_from" = "CDS_start",
    "cds_strand" = "strand",
    "cds_to" = "CDS_end",
    "cds_locations" = "CDS_ranges"
  ) %>%
  relocate(chr, .before = nc_accession) %>%
  relocate(gene, .after = nc_accession) %>%
  relocate(cds_locations, .after = cds_to)

```

# Process CDS data from T2T-CHM13 for the sgRNA library refinement pipeline

```{r}

cds_gene$cds_from <- as.integer(cds_gene$cds_from)
cds_gene$cds_to <- as.integer(cds_gene$cds_to)

list_chromosome <- c(1:22, 'X', 'Y')

cds_gene <- cds_gene %>%
    filter(!is.na(cds_from), !is.na(cds_to))

cds_gene_exon <- cds_gene %>%
    tidyr::unnest(cds_locations) %>%
    group_by(gene, gene_id) %>%
    mutate(exon_code = ifelse(cds_strand == "+", 1:dplyr::n(), dplyr::n():1)) %>%
    ungroup() %>%
    dplyr::mutate(cds_start = str_extract(cds_locations, "^[0-9]+") %>% as.integer,
                  cds_end = str_extract(cds_locations, "[0-9]+$") %>% as.integer) %>%
    dplyr::select(gene, gene_id, chr, start = cds_start, end = cds_end, strand = cds_strand,
                    gene_start = cds_from, gene_end = cds_to, exon_code)

genome_info <- GenomeInfoDb::Seqinfo(genome = "T2T-CHM13v2.0")[list_chromosome]

gene_annot_granges <- cds_gene_exon %>% 
      GenomicRanges::makeGRangesFromDataFrame(seqinfo = genome_info, 
                                              keep.extra.columns = T, 
                                              na.rm = TRUE)
saveRDS(gene_annot_granges, file = "T2T-CHM13v2.0_gene_annot_granges.rds")

```

