---
title: "Compare sgRNA Library performance"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}

library(tidyverse)
library(VennDiagram)
library(ggvenn)

source('sgRNA_library_refining_functions.R')

```

# Import sgRNA library data

```{r}
# import sample sgRNA library

import_sgrna_library <- function(data_dir) {
  library <- read_tsv(data_dir, col_names = FALSE)
  colnames(library) <- c('sgRNA', 'spacer', 'gene')
  return(library)
}

brunello <- import_sgrna_library('./public_crispr_library/broadgpp-brunello-library-contents.tsv')
toronto <- import_sgrna_library('./public_crispr_library/tkov3_guide_sequence.tsv')
cellecta_M1 <- import_sgrna_library('./cellecta_library/sample_M1_150K.tsv')
cellecta_M2 <- import_sgrna_library('./cellecta_library/sample_M2_150K.tsv')
cellecta_M3 <- import_sgrna_library('./cellecta_library/sample_M3_150K.tsv')

# combined cellecta and remove duplicates
cellecta <- rbind(cellecta_M1, cellecta_M2, cellecta_M3)
cellecta <- cellecta %>%
  arrange(sgRNA) %>%
  distinct()

```

# A venn diagram function to compare the similarity between three libraries

```{r}

venn_diagram <- function(set1, set2, name_set1, name_set2, color1, color2, name_venn) {
  only_set1 <- length(setdiff(set1, set2))
  only_set2 <- length(setdiff(set2, set1))
  both_sets  <- length(intersect(set1, set2))

  totalset1 <- only_set1 + both_sets
  totalset2 <- only_set2 + both_sets

  pct_set1 <- round((only_set1 / totalset1) * 100, 2)
  pct_set2 <- round((only_set2 / totalset2) * 100, 2)
  pct_overlap_set1 <- round((both_sets / totalset1) * 100, 2)
  pct_overlap_set2 <- round((both_sets / totalset2) * 100, 2)

  sgRNA_similarity <- venn.diagram(
        x = list(set1, set2),
        category.names = c(name_set1, name_set2),
        filename = NULL,  
        fill = c(color1, color2),
        alpha = 0.5,
        cex = 0,  # Set text size to 0 to hide numbers
        cat.cex = 0.75,  # Set category label size
        cat.pos = c(-2, 2),  # Adjust category label positions
        cat.dist = c(0.05, 0.05),  # Adjust category label distances
        lwd = 2
  )

  png(filename = name_venn, width = 1250, height = 1250, res = 300)
  grid.newpage()
  grid.draw(sgRNA_similarity)

  grid.text(label = paste0(only_set2, '\n', pct_set1, '%'), x = 0.25, y = 0.50, gp = gpar(fontsize = 10))
  grid.text(label = paste0(only_set1, '\n', pct_set2, '%'), x = 0.75, y = 0.50, gp = gpar(fontsize = 10))
  grid.text(label = paste0(pct_overlap_set1, '%', '\nor\n', pct_overlap_set2, '%'), x = 2.5, y = 0.50, gp = gpar(fontsize = 7))

  dev.off()
}


```

# compare the similarity of all sequences between three libraries

```{r}
venn_diagram(brunello$spacer, toronto$spacer, "Brunello \n 77K sgRNAs", "Toronto V3 \n 71K sgRNAs", 'powderblue', 'mistyrose', 'VennDiagram_sgRNAs_brunello_toronto.png')
venn_diagram(brunello$spacer, cellecta$spacer, "Brunello \n 77K sgRNAs", "Cellecta \n 150K sgRNAs", 'powderblue', 'lightgreen', 'VennDiagram_sgRNAs_brunello_cellecta.png') 
venn_diagram(toronto$spacer, cellecta$spacer, "Toronto V3 \n 77K sgRNAs", "Cellecta \n 150K sgRNAs", 'mistyrose', 'lightgreen', 'VennDiagram_sgRNAs_toronto_cellecta.png') 
```

# import alignment data from all libraries

```{r}

brunello_alignment <- read_csv('./object_intermediate/broadgpp-brunello-library-contents_aln.csv')
toronto_alignment <- read_csv('./object_intermediate/tkov3_guide_sequence_aln.csv')
cellecta_M1_alignment <- read_csv('./object_intermediate/sample_M1_150K_aln.csv')
cellecta_M2_alignment <- read_csv('./object_intermediate/sample_M2_150K_aln.csv')
cellecta_M3_alignment <- read_csv('./object_intermediate/sample_M3_150K_aln.csv')

# combine all cellecta alignment
cellecta_alignment <- rbind(cellecta_M1_alignment, cellecta_M2_alignment, cellecta_M3_alignment)
cellecta_alignment <- cellecta_alignment %>%
  arrange(sgRNA) %>%
  distinct()

# keep alignment only for normal chromosome
brunello_alignment <- alignment_normal_chr(brunello_alignment)
toronto_alignment <- alignment_normal_chr(toronto_alignment)
cellecta_alignment <- alignment_normal_chr(cellecta_alignment)

```

# detect multi-target sgRNAs

```{r}

brunello_multi_target <- multi_target_sgrna(brunello_alignment)
brunello_single_mismatch <- single_mismatch_sgrna(brunello_alignment)
brunello_pam_mismatch <- pam_distal_double_mismatch(brunello_alignment)

brunello_off_target <- unique(union(brunello_pam_mismatch$list_guides, unique(union(brunello_multi_target$list_guides, brunello_single_mismatch$list_guides))))

toronto_multi_target <- multi_target_sgrna(toronto_alignment)
toronto_single_mismatch <- single_mismatch_sgrna(toronto_alignment)
toronto_pam_mismatch <- pam_distal_double_mismatch(toronto_alignment)

toronto_off_target <- unique(union(toronto_pam_mismatch$list_guides, unique(union(toronto_multi_target$list_guides, toronto_single_mismatch$list_guides))))

cellecta_multi_target <- multi_target_sgrna(cellecta_alignment)
cellecta_single_mismatch <- single_mismatch_sgrna(cellecta_alignment)
cellecta_pam_mismatch <- pam_distal_double_mismatch(cellecta_alignment)

cellecta_off_target <- unique(union(cellecta_pam_mismatch$list_guides, unique(union(cellecta_multi_target$list_guides, cellecta_single_mismatch$list_guides))))

brunello_off_target_spacer <- brunello %>%
  filter(sgRNA %in% brunello_off_target)

toronto_off_target_spacer <- toronto %>%
  filter(sgRNA %in% toronto_off_target)

cellecta_off_target_spacer <- cellecta %>%
  filter(sgRNA %in% cellecta_off_target)

off_target_venn <- venn.diagram(
    x = list(
        cellecta_off_target_spacer$spacer,
        brunello_off_target_spacer$spacer
    ),
    category.names = c("Cellecta \n 150K sgRNAs", "Brunello \n 77K sgRNAs"),
    filename = NULL,
    output = TRUE,
    fill = c("lightgreen", "powderblue"),
    alpha = 0.5,
    cex = 2.5,
    cat.cex = 2,
    cat.pos = c(-20, 20),
    cat.dist = c(0.1, 0.1)
)

ggsave(
    path = './',
    width = 10,
    height = 10,
    dpi = 300,
    plot = off_target_venn,
    filename = paste('VennDiagram_off_target_cellecta_brunello.png', sep = ''))

```

# Analysis of multi-target sgRNAs

```{r}

# identification of sgRNAs

brunello_multi_target <- multi_target_sgrna(brunello_alignment)
toronto_multi_target <- multi_target_sgrna(toronto_alignment)
cellecta_multi_target <- multi_target_sgrna(cellecta_alignment)

brunello_single_mismatch <- single_mismatch_sgrna(brunello_alignment)
toronto_single_mismatch <- single_mismatch_sgrna(brunello_alignment)
cellecta_single_mismatch <- single_mismatch_sgrna(brunello_alignment)

brunello_single_mismatch_list_guides <- setdiff(brunello_single_mismatch$list_guides, brunello_multi_target$list_guides)
toronto_single_mismatch_list_guides <- setdiff(toronto_single_mismatch$list_guides, toronto_multi_target$list_guides)
cellecta_single_mismatch_list_guides <- setdiff(cellecta_single_mismatch$list_guides, cellecta_multi_target$list_guides)

# alignment data

brunello_alignment_multi_target <- brunello_alignment %>% filter(sgRNA %in% brunello_multi_target$list_guides)
toronto_alignment_multi_target <- toronto_alignment %>% filter(sgRNA %in% toronto_multi_target$list_guides)
cellecta_alignment_multi_target <- cellecta_alignment %>% filter(sgRNA %in% cellecta_multi_target$list_guides)

brunello_alignment_single_mismatch <- brunello_alignment %>% filter(sgRNA %in% brunello_single_mismatch_list_guides)
toronto_alignment_single_mismatch <- toronto_alignment %>% filter(sgRNA %in% toronto_single_mismatch_list_guides)
cellecta_alignment_single_mismatch <- cellecta_alignment %>% filter(sgRNA %in% cellecta_single_mismatch_list_guides)

```

# Calculate the number of multi-target sgRNAs targeting other genomic locations and protein-coding genes

```{r}

annotate_off_target <- function(alignment_data) {
  ccds <- read_ccds_data('CCDS_20250112.txt')
  transformed_ccds <- transform_gene_annotation_ccds(ccds)
  modified_alignment <- add_cut_pos_pam_pos(alignment_data)
  off_target_granges <- make_granges_from_alignment_data(modified_alignment)
  gene_df_overlap <- find_overlaps_gene_annotation_and_alignment(off_target_granges, transformed_ccds$gene_annot_granges_df)

  return(gene_df_overlap)
    
}

off_target_analysis <- function(alignment_data, num_mismatch, library, library_name){
  gene_df <- annotate_off_target(alignment_data)
  gene_df <- gene_df %>%
    dplyr::rename('actual_chr' = 'chr',
                  'actual_gene' = 'gene')

  alignment_data <- add_cut_pos_pam_pos(alignment_data)
  alignment_data <- alignment_data %>%
    select(unique_aln_id, sgRNA, spacer, protospacer, gene, chr, n_mismatches, cut_pos) %>% 
    dplyr::rename('sgrna' = 'sgRNA')

  alignment_data_gene_df <- alignment_data %>%
    left_join(gene_df, by = join_by(unique_aln_id, sgrna, spacer, protospacer, cut_pos)) %>%
    select(unique_aln_id, sgrna, spacer, protospacer, gene, chr, n_mismatches, cut_pos, actual_chr, actual_gene) %>%
    filter(n_mismatches == num_mismatch) %>%
    mutate(target = case_when(gene == actual_gene ~ 'actual gene',
                            gene != actual_gene ~ 'protein-coding genes',
                            is.na(actual_gene) ~ 'random genomic loci')) %>%
    filter(target != 'actual gene')

  summary_target_df <- alignment_data_gene_df %>%
    select(sgrna, target) %>%
    group_by(sgrna) %>%
    mutate(target = ifelse(n() > 1, 'random genomic loci or protein-coding genes', target)) %>%
    ungroup() %>%
    distinct()
  
  calculate_freq_table <- summary_target_df %>%
    select(target) %>%
    count(target) %>%
    dplyr::rename('count_sgrna' = 'n') %>%
    arrange(desc(count_sgrna)) %>%
    mutate(percent = round(count_sgrna/length(library$sgRNA)*100 ,2),
           library = library_name)

  return(list(
    aln_data_df = alignment_data_gene_df,
    summary_df = summary_target_df, 
    freq_table = calculate_freq_table))

}

brunello_multi_target_analysis <- off_target_analysis(brunello_alignment_multi_target, 0, brunello, 'brunello')
toronto_multi_target_analysis <- off_target_analysis(toronto_alignment_multi_target, 0, toronto, 'toronto')
cellecta_multi_target_analysis <- off_target_analysis(cellecta_alignment_multi_target, 0, cellecta, 'cellecta')

brunello_single_mismatch_analysis <- off_target_analysis(brunello_alignment_single_mismatch, 1, brunello, 'brunello')
toronto_single_mismatch_analysis <- off_target_analysis(toronto_alignment_single_mismatch, 1, toronto, 'toronto')
cellecta_single_mismatch_analysis <- off_target_analysis(cellecta_alignment_single_mismatch, 1, cellecta, 'cellecta')

```

# Frequency table and visualization

```{r}

table_multi_target <- rbind(brunello_multi_target_analysis$freq_table,
                            toronto_multi_target_analysis$freq_table,
                            cellecta_multi_target_analysis$freq_table)
table_multi_target <- table_multi_target %>%
  arrange(target)

table_single_mismatch <- rbind(brunello_single_mismatch_analysis$freq_table,
                              toronto_single_mismatch_analysis$freq_table,
                              cellecta_single_mismatch_analysis$freq_table)
table_single_mismatch <- table_single_mismatch %>%
  arrange(target)


# Grouped bar chart
grouped_barplot <- ggplot(table_multi_target, aes(x = factor(library, levels = c('cellecta', 'brunello', 'toronto')), y = percent, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Percentage of sgRNAs (%)") +
  scale_x_discrete(labels = c('brunello' = 'Brunello \n 77K sgRNAs',
                'cellecta' = 'Cellecta \n 150K sgRNAs',
                'toronto' = 'Toronto \nKnockout V3 \n 71K sgRNAs')) +
  scale_fill_manual(values = c("random genomic loci" = "#B3589A",
                              "protein-coding genes" = "#E7F7D5",
                              "random genomic loci or protein-coding genes" = "#ff9028")) +
  theme(axis.text.x = element_text(size = 14),
      axis.text.y = element_text(size = 14),
      axis.title.y = element_text(size = 16, margin = margin(r = 10)),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12)) +
  geom_text(aes(label = paste0(percent, "%")), position = position_dodge(width = 0.8), vjust = -0.5, size = 4)  +
  ylim(0, 3)

grouped_barplot

ggsave(
    path = './',
    width = 15,
    height = 15,
    dpi = 300,
    plot = grouped_barplot,
    filename = paste('grouped_barplot_multi_target_2.png', sep = ''))

```

# Paralog pair analysis

```{r}

# are multi-target sgRNAs more likely to target paralog genes ?

predicted_paralog_dataset <- read_csv('barbara_paralog_dataset.csv')
predicted_paralog_dataset <- predicted_paralog_dataset %>%
  mutate(sorted_gene_pair_2 = paste(A2, '_', A1, sep = '')) %>%
  relocate(sorted_gene_pair_2, .after = sorted_gene_pair)

brunello_off_target_paralog <- brunello_multi_target_analysis$aln_data_df %>%
  mutate(gene_pair = paste(gene, actual_gene, sep = '_')) %>%
  filter(gene_pair %in% predicted_paralog_dataset$sorted_gene_pair)

brunello_off_target_not_paralog <- brunello_multi_target_analysis$aln_data_df %>%
  mutate(gene_pair = paste(gene, actual_gene, sep = '_')) %>%
  filter(!gene_pair %in% predicted_paralog_dataset$sorted_gene_pair) %>%
  filter(!sgrna %in% brunello_off_target_paralog$sgrna)

# intersection <- intersect(unique(brunello_off_target_paralog$sgrna), unique(brunello_off_target_not_paralog$sgrna))

common_paralog_brunello <- brunello_alignment %>%
  filter(!sgRNA %in% brunello_multi_target$list_guides, 
          (gene %in% predicted_paralog_dataset$A1) & (gene %in% predicted_paralog_dataset$A2))

head(brunello_alignment, 100)

test <- brunello_off_target_paralog %>% filter(sgrna %in% intersection)
test_2 <- brunello_off_target_not_paralog %>% filter(sgrna %in% intersection)

n_distinct(brunello_off_target_paralog$sgrna)
n_distinct(brunello_off_target_not_paralog$sgrna)
n_distinct(common_paralog_brunello$sgRNA)

1041+1374

fischer_exact_test_for_off_target <- function(library_multi_target_analysis_aligned_data, original_alignment_data, off_target_alignment_data, paralog_dataset) {
  
  predicted_paralog_dataset <- read_csv(paralog_dataset)
  library_aln_data_df <- library_multi_target_analysis_aligned_data$aln_data_df

  off_target_paralog <- library_aln_data_df %>%
    mutate(gene_pair = paste(gene, actual_gene, sep = '_')) %>%
    filter(gene_pair %in% predicted_paralog_dataset$sorted_gene_pair)

  off_target_not_paralog <- library_aln_data_df %>%
    mutate(gene_pair = paste(gene, actual_gene, sep = '_')) %>%
    filter(!gene_pair %in% predicted_paralog_dataset$sorted_gene_pair) %>%
    filter(!sgrna %in% off_target_paralog$sgrna)

  common_paralog <- original_alignment_data %>%
    filter(!sgRNA %in% off_target_alignment_data$list_guides, 
          (gene %in% predicted_paralog_dataset$A1) & (gene %in% predicted_paralog_dataset$A2))

  common_not_paralog <- original_alignment_data %>%
    filter(!sgRNA %in% off_target_alignment_data$list_guides, 
          (!gene %in% predicted_paralog_dataset$A1) | (!gene %in% predicted_paralog_dataset$A2),
          !sgRNA %in% common_paralog$sgRNA)

  print(paste('Multi-target sgRNAs targeting paralog genes:', n_distinct(off_target_paralog$sgrna)))
  print(paste('Multi-target sgRNAs not targeting paralog genes:', n_distinct(off_target_not_paralog$sgrna)))
  print(paste('Non multi-target sgRNAs targeting paralog genes:', n_distinct(common_paralog$sgRNA)))
  print(paste('Non multi-target sgRNAs not targeting paralog genes:', n_distinct(common_not_paralog$sgRNA)))

  # Create the contingency table
  contingency_table <- matrix(
    c(
      length(unique(off_target_paralog$sgrna)),  # Multi-target sgRNAs targeting paralog genes
      length(unique(off_target_not_paralog$sgrna)),  # Multi-target sgRNAs not targeting paralog genes
      length(unique(common_paralog$sgRNA)),  # Non-multi-target sgRNAs targeting paralog genes
      length(unique(common_not_paralog$sgRNA))  # Non-multi-target sgRNAs not targeting paralog genes
    ),
    nrow = 2
  )

  print(contingency_table)

  # Check if the total matches the unique sgRNAs in alignment_data
  total_contingency_table <- sum(contingency_table)
  total_unique_sgrna <- length(unique(original_alignment_data$sgRNA))

  print(paste('total sgRNA count in contingency table:', total_contingency_table))
  print(paste('total original number of sgRNA:', total_unique_sgrna))

  if (total_contingency_table == total_unique_sgrna) {
    print("The sum of the contingency table matches the total unique sgRNAs in alignment_data.")
  } else {
    print("Mismatch detected between the contingency table sum and the total unique sgRNAs in alignment_data.")
  }

  # Perform Fisher's test
  fischer_library_multi_target <- fisher.test(contingency_table)
  print(fischer_library_multi_target)

  return(list(off_target_paralog_df = off_target_paralog,
              off_target_not_paralog_df = off_target_not_paralog,
              common_paralog_df = common_paralog, 
              common_not_paralog_df = common_not_paralog))

}


dataset_brunello_fischer <- fischer_exact_test_for_off_target(brunello_multi_target_analysis, 
                                                              brunello_alignment, 
                                                              brunello_multi_target, 
                                                              'barbara_paralog_dataset.csv')

dataset_toronto_fischer <- fischer_exact_test_for_off_target(toronto_multi_target_analysis, 
                                                              toronto_alignment, 
                                                              toronto_multi_target, 
                                                              'barbara_paralog_dataset.csv')

dataset_cellecta_fischer <- fischer_exact_test_for_off_target(cellecta_multi_target_analysis, 
                                                              cellecta_alignment, 
                                                              cellecta_multi_target, 
                                                              'barbara_paralog_dataset.csv')



# WHY THIS RESULT IS DIFFERENT IF COMPARED TO THE ONE IN THE FUNCTION FUCK

?fisher.test

contingency_table_test <- matrix(
    c(1041, 1879, 23561, 50794),
    nrow = 2)
contingency_table_test
fisher.test(contingency_table_test)

n_distinct(toronto$gene)

```


# brunello_aln_data_df <- brunello_multi_target_analysis$aln_data_df
# multi-target sgRNAs targeting genes with paralog pairs
# 1041
# multi-target sgRNAs targeting genes that are not paralog pairs
# 1879
# total multi-target --> 2920
# the rest of sgRNA (total sgRNA - multi-target sgRNA targeting paralog)
# 23561
# 50794
# 23561+50794+1041+1879
