---
title: "annotation of multi-target sgRNAs"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(tidyverse)
library(VennDiagram)
library(ggvenn)

source('../sgRNA_library_refinement_pipeline_functions.R')
```

# List of functions

```{r}
# import sgRNA library
import_sgrna_library <- function(data_dir) {
  library <- read_tsv(data_dir, col_names = FALSE)
  colnames(library) <- c('sgRNA', 'spacer', 'gene')
  return(library)
}

# import alignment data and keep alignment for normal chromosome only
import_sgrna_library_alignment <- function(data_dir) {
  library <- read_csv(data_dir)
  aln_normal_chr <- alignment_normal_chr(library)
  return(aln_normal_chr)
}

# get list off target guides
list_off_target_guides <- function(alignment_data) {
  # start with multi-target sgRNA
  multi_target_library <- multi_target_sgrna(alignment_data)
  
  # removal of multi-target sgRNA from the library
  alignment_data_2 <- alignment_data %>%
    filter(!sgRNA %in% multi_target_library$list_guides)
  
  # continue with single mismatch sgRNA
  single_mismatch_library <- single_mismatch_sgrna(alignment_data_2)
  
  # removal of single mismatch sgRNA from the library
  alignment_data_3 <- alignment_data_2 %>%
    filter(!sgRNA %in% single_mismatch_library$list_guides)
  
  # continue with pam double mismatch sgRNA
  pam_double_mismatch_library <- pam_distal_double_mismatch(alignment_data_3)
  
  list_off_target_guides <- unique(union(multi_target_library$list_guides, unique(union(single_mismatch_library$list_guides, pam_double_mismatch_library$list_guides))))
  
  return(list(
    multi_target_guides = unique(multi_target_library$list_guides),
    single_mismatch_guides = unique(single_mismatch_library$list_guides), 
    pam_double_mismatch_guides = unique(pam_double_mismatch_library$list_guides),
    all_off_target_guides = list_off_target_guides))
}

# annotate off-target sgRNAs
annotate_off_target <- function(alignment_data) {
  ccds <- read_ccds_data('../CCDS_20250112.txt')
  transformed_ccds <- transform_gene_annotation_ccds(ccds)
  modified_alignment <- add_cut_pos_pam_pos(alignment_data)
  off_target_granges <- make_granges_from_alignment_data(modified_alignment)
  gene_df_overlap <- find_overlaps_gene_annotation_and_alignment(off_target_granges, transformed_ccds$gene_annot_granges_df)
  return(gene_df_overlap)
}

# off target annotation (can be multi-target, single-mismatch, or pam-distal double mismatches) used to annotate whether the sgRNA target random genomic loci, protein-coding genes or intended target gene

off_target_annotation <- function(alignment_data, num_mismatch, library, library_name){
  gene_df <- annotate_off_target(alignment_data)
  gene_df <- gene_df %>%
    dplyr::rename('actual_chr' = 'chr',
                  'actual_gene' = 'gene')

  alignment_data <- add_cut_pos_pam_pos(alignment_data)
  alignment_data <- alignment_data %>%
    select(unique_aln_id, sgRNA, spacer, protospacer, gene, chr, n_mismatches, cut_pos) %>%
    dplyr::rename('sgrna' = 'sgRNA')
  
  # one version of alignment data for paralog analysis (you will need the actual gene to detect which multi-target sgRNA targeting only two protein-coding genes - one is the actual gene, the other one is the paralog pairs)
  alignment_data_gene_df_1 <- alignment_data %>%
    left_join(gene_df, by = join_by(unique_aln_id, sgrna, spacer, protospacer, cut_pos)) %>%
    select(unique_aln_id, sgrna, spacer, protospacer, gene, chr, n_mismatches, cut_pos, actual_chr, actual_gene) %>%
    filter(n_mismatches == num_mismatch) %>%
    mutate(target = case_when(gene == actual_gene ~ 'actual gene',
                              gene != actual_gene ~ 'protein-coding genes',
                              is.na(actual_gene) ~ 'other genomic loci'))
  
  # 2nd version of alignment data for frequency table and histogram plot
  alignment_data_gene_df_2 <- alignment_data_gene_df_1 %>%
    filter(target != 'actual gene')
  
  summary_target_df <- alignment_data_gene_df_2 %>%
    select(sgrna, target) %>%
    group_by(sgrna) %>%
    mutate(target = ifelse(n() > 1, 'other genomic loci or protein-coding genes', target)) %>%
    ungroup() %>%
    distinct()
  
  calculate_freq_table <- summary_target_df %>%
    select(target) %>%
    count(target) %>%
    dplyr::rename('count_sgrna' = 'n') %>%
    arrange(desc(count_sgrna)) %>%
    mutate(percent = round(count_sgrna/length(library$sgRNA)*100 ,2),
           library = library_name)

  return(list(
    aln_data_df_1 = alignment_data_gene_df_1,
    aln_data_df_2 = alignment_data_gene_df_2,
    summary_df = summary_target_df, 
    freq_table = calculate_freq_table))
}


```

# import alignment data from all libraries and import sgRNA libraries

```{r}
hgnc <- read_tsv(url('https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt'))

brunello <- import_sgrna_library('../public_crispr_library/broadgpp-brunello-library-contents.tsv')
toronto <- import_sgrna_library('../public_crispr_library/tkov3_guide_sequence.tsv')
yusa <- import_sgrna_library('../public_crispr_library/yusa_hcrispr_ko_grnas.tsv')
avana <- import_sgrna_library('../public_crispr_library/avana_library.tsv')

brunello_alignment <- import_sgrna_library_alignment('../object_intermediate/broadgpp-brunello-library-contents_aln.csv')
toronto_alignment <- import_sgrna_library_alignment('../object_intermediate/tkov3_guide_sequence_aln.csv')
yusa_alignment <- import_sgrna_library_alignment('../object_intermediate/yusa_hcrispr_ko_grnas_aln.csv')
avana_alignment <- import_sgrna_library_alignment('../object_intermediate/avana_library_aln.csv')

# cellecta_M1_alignment <- import_sgrna_library_alignment('./object_intermediate/sample_M1_150K_aln.csv')
# cellecta_M2_alignment <- import_sgrna_library_alignment('./object_intermediate/sample_M2_150K_aln.csv')
# cellecta_M3_alignment <- import_sgrna_library_alignment('./object_intermediate/sample_M3_150K_aln.csv')
# 
# cellecta_alignment <- rbind(cellecta_M1_alignment, cellecta_M2_alignment, cellecta_M3_alignment)
```

# detect all type of off-target sgRNAs from all libraries

```{r}

brunello_off_target <- list_off_target_guides(brunello_alignment)
toronto_off_target <- list_off_target_guides(toronto_alignment)
yusa_off_target <- list_off_target_guides(yusa_alignment)
avana_off_target <- list_off_target_guides(avana_alignment)

```

# annotation of multi-target sgRNAs

```{r}
# alignment data

brunello_alignment_multi_target <- brunello_alignment %>% filter(sgRNA %in% brunello_off_target$multi_target_guides)
toronto_alignment_multi_target <- toronto_alignment %>% filter(sgRNA %in% toronto_off_target$multi_target_guides)
yusa_alignment_multi_target <- yusa_alignment %>% filter(sgRNA %in% yusa_off_target$multi_target_guides)
avana_alignment_multi_target <- avana_alignment %>% filter(sgRNA %in% avana_off_target$multi_target_guides)

brunello_alignment_single_mismatch <- brunello_alignment %>% filter(sgRNA %in% brunello_off_target$single_mismatch_guides)
toronto_alignment_single_mismatch <- toronto_alignment %>% filter(sgRNA %in% toronto_off_target$single_mismatch_guides)
yusa_alignment_single_mismatch <- yusa_alignment %>% filter(sgRNA %in% yusa_off_target$single_mismatch_guides)
avana_alignment_single_mismatch <- avana_alignment %>% filter(sgRNA %in% avana_off_target$single_mismatch_guides)

# for cellecta
# cellecta_multi_target <- multi_target_sgrna(cellecta_alignment)
# cellecta_single_mismatch <- single_mismatch_sgrna(brunello_alignment)
# cellecta_single_mismatch_list_guides <- setdiff(cellecta_single_mismatch$list_guides, cellecta_multi_target$list_guides)
# cellecta_alignment_multi_target <- cellecta_alignment %>% filter(sgRNA %in% cellecta_off$list_guides)
# cellecta_alignment_single_mismatch <- cellecta_alignment %>% filter(sgRNA %in% cellecta_single_mismatch_list_guides)

```

# Calculate the number of multi-target sgRNAs targeting other genomic locations and protein-coding genes

```{r}

brunello_multi_target_annotation <- off_target_annotation(brunello_alignment_multi_target, 0, brunello, 'brunello')
toronto_multi_target_annotation <- off_target_annotation(toronto_alignment_multi_target, 0, toronto, 'toronto')
yusa_multi_target_annotation <- off_target_annotation(yusa_alignment_multi_target, 0, yusa, 'yusa')
avana_multi_target_annotation <- off_target_annotation(avana_alignment_multi_target, 0, avana, 'avana')

brunello_single_mismatch_annotation <- off_target_annotation(brunello_alignment_single_mismatch, 1, brunello, 'brunello')
toronto_single_mismatch_annotation <- off_target_annotation(toronto_alignment_single_mismatch, 1, toronto, 'toronto')
yusa_single_mismatch_annotation <- off_target_annotation(yusa_alignment_single_mismatch, 1, yusa, 'yusa')
avana_single_mismatch_annotation <- off_target_annotation(avana_alignment_single_mismatch, 1, avana, 'avana')

# for cellecta
# cellecta_multi_target_annotation <- off_target_annotation(cellecta_alignment_multi_target, 0, cellecta, 'cellecta')
# cellecta_single_mismatch_annotation <- off_target_annotation(cellecta_alignment_single_mismatch, 1, cellecta, 'cellecta')
```

### What is percentage of sgRNAS that target multiple genomic sites perfectly?
### What is percentage of sgRNAS that target multiple protein-coding genes perfectly?
### What is percentage of sgRNAS that target multiple genomic sites with mismatches?
### What is percentage of sgRNAS that target multiple protein-coding genes with mismatches?

```{r}

table_multi_target <- rbind(brunello_multi_target_annotation$freq_table,
                            toronto_multi_target_annotation$freq_table,
                            yusa_multi_target_annotation$freq_table,
                            avana_multi_target_annotation$freq_table)
table_multi_target <- table_multi_target %>%
  arrange(target)

table_single_mismatch <- rbind(brunello_single_mismatch_annotation$freq_table,
                              toronto_single_mismatch_annotation$freq_table,
                              yusa_single_mismatch_annotation$freq_table,
                              avana_single_mismatch_annotation$freq_table)
table_single_mismatch <- table_single_mismatch %>%
  arrange(target)


# Grouped bar chart
grouped_barplot <- ggplot(table_single_mismatch, aes(x = factor(library, levels = c('brunello', 'toronto', 'yusa', 'avana')), y = percent, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Percentage of single mismatch sgRNAs (%)") +
  scale_x_discrete(labels = c('brunello' = 'Brunello \n 77K sgRNAs',
                'yusa' = 'Yusa \n 90K sgRNAs',
                'toronto' = 'Toronto \nKnockout V3 \n 71K sgRNAs',
                'avana' = 'Avana 78K sgRNAs')) +
  scale_fill_manual(values = c("other genomic loci" = "#B3589A",
                              "protein-coding genes" = "#E7F7D5",
                              "other genomic loci or protein-coding genes" = "#ff9028")) +
  theme(axis.text.x = element_text(size = 14),
      axis.text.y = element_text(size = 14),
      axis.title.x = element_text(size = 16, margin = margin(r = 10)),
      axis.title.y = element_text(size = 16, margin = margin(r = 10)),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12)) +
  geom_text(aes(label = paste0(percent, "%")), position = position_dodge(width = 0.8), vjust = -0.5, size = 4)  +
  ylim(0, 3)

grouped_barplot

ggsave(
    path = './',
    width = 15,
    height = 15,
    dpi = 300,
    plot = grouped_barplot,
    filename = paste('grouped_barplot_single_mismatch.png', sep = ''))


```

### Are multitargeting effects more likely to target pairs of paralogs?

```{r}

# list paralog gene pairs from barbara's dataset
list_paralog_gene_pairs <- function(dataset) {
  predicted_paralog_dataset <- read_csv(dataset)
  gene_pair_1 <- predicted_paralog_dataset$sorted_gene_pair
  gene_pair_2 <- predicted_paralog_dataset$sorted_gene_pair_2
  
  gene_pair <- unique(union(gene_pair_1, gene_pair_2))
  return(gene_pair)
}

# pick multi-target alignment strictly with 2 protein-coding genes only

find_multi_target_paralog <- function(alignment_gene_df_1, paralog_dataset, how_many_target_paralog) {
  
  aln_multi_target_paralog <- alignment_gene_df_1 %>%
    drop_na() %>%
    mutate(target = case_when(paste(gene, actual_gene, sep ='_') %in% predicted_paralog_pairs ~ 'targeting paralog gene',
                              gene == actual_gene ~ 'actual gene',
                              TRUE ~ 'other protein-coding gene')) %>%
    group_by(sgrna) %>%
    filter(n() == how_many_target_paralog,
           all(c("targeting paralog gene", "actual gene") %in% target),
           n_distinct(target) == 2) %>% 
    ungroup()
  
  list_not_paralog_guides <- setdiff(unique(alignment_gene_df_1$sgrna), unique(aln_multi_target_paralog$sgrna))
  
  return(list(aln_multi_target_paralog_df = aln_multi_target_paralog,
              multi_target_paralog_guides = unique(aln_multi_target_paralog$sgrna),
              multi_target_not_paralog_guides = unique(list_not_paralog_guides)))
  
}

find_not_multi_target_paralog <- function(sgrna_library, alignment_gene_df_1, paralog_dataset) {
  
  # read paralog dataset
  paralog <- read_csv(paralog_dataset)
  
  # non multi-target sgRNA in the library
  sgrna_library_not_multi_target <- sgrna_library %>%
    filter(!sgRNA %in% alignment_gene_df_1$sgrna)
  
  # normal sgRNA targeting paralog
  sgrna_library_paralog <- sgrna_library_not_multi_target %>%
    filter(gene %in% unique(paralog$A2))
  
  # normal sgRNA not targeting paralog
  sgrna_library_not_paralog <- setdiff(unique(sgrna_library_not_multi_target$sgRNA), unique(sgrna_library_paralog$sgRNA))
  
  return (list(normal_sgrna_paralog = unique(sgrna_library_paralog$sgRNA),
               normal_sgrna_not_paralog = sgrna_library_not_paralog))
  
}

fisher_multi_target_paralog <- function(off_target_paralog_guides, off_target_not_paralog_guides, normal_paralog_guides, normal_not_paralog_guides, sgrna_library) {

  print(paste('Multi-target sgRNAs targeting paralog genes:', length(off_target_paralog_guides)))
  print(paste('Multi-target sgRNAs not targeting paralog genes:', length(off_target_not_paralog_guides)))
  print(paste('Non multi-target sgRNAs targeting paralog genes:', length(normal_paralog_guides)))
  print(paste('Non multi-target sgRNAs not targeting paralog genes:', length(normal_not_paralog_guides)))

  # Create the contingency table
  contingency_table <- matrix(
    c(
      length(off_target_paralog_guides),  # Multi-target sgRNAs targeting paralog genes
      length(off_target_not_paralog_guides),  # Multi-target sgRNAs not targeting paralog genes
      length(normal_paralog_guides),  # Non-multi-target sgRNAs targeting paralog genes
      length(normal_not_paralog_guides)  # Non-multi-target sgRNAs not targeting paralog genes
    ),
    nrow = 2
  )

  print(contingency_table)

  # Check if the total matches the unique sgRNAs in alignment_data
  total_contingency_table <- sum(contingency_table)
  total_unique_sgrna <- length(unique(sgrna_library$sgRNA))

  print(paste('total sgRNA count in contingency table:', total_contingency_table))
  print(paste('total original number of sgRNA:', total_unique_sgrna))

  if (total_contingency_table == total_unique_sgrna) {
    print("The sum of the contingency table matches the total unique sgRNAs in alignment_data.")
  } else {
    print("Mismatch detected between the contingency table sum and the total unique sgRNAs in alignment_data.")
  }

  # Perform Fisher's test
  fischer_library_multi_target <- fisher.test(contingency_table)
  print(fischer_library_multi_target)

}

predicted_paralog_pairs <- list_paralog_gene_pairs('../barbara_36K_paralog_pairs.csv')

# Brunello paralog analysis
brunello_multi_target_paralog <- find_multi_target_paralog(brunello_multi_target_annotation$aln_data_df_1, predicted_paralog_pairs, 3)

brunello_not_paralog <- find_not_multi_target_paralog(brunello, brunello_multi_target_annotation$aln_data_df_1, '../barbara_36K_paralog_pairs.csv')

dataset_brunello_fischer <- fisher_multi_target_paralog(brunello_multi_target_paralog$multi_target_paralog_guides, 
                                                        brunello_multi_target_paralog$multi_target_not_paralog_guides, 
                                                        brunello_not_paralog$normal_sgrna_paralog, 
                                                        brunello_not_paralog$normal_sgrna_not_paralog,
                                                        brunello)

# Toronto paralog analysis
toronto_multi_target_paralog <- find_multi_target_paralog(toronto_multi_target_annotation$aln_data_df_1, predicted_paralog_pairs, 3)

toronto_not_paralog <- find_not_multi_target_paralog(toronto, toronto_multi_target_annotation$aln_data_df_1, '../barbara_36K_paralog_pairs.csv')

dataset_toronto_fischer <- fisher_multi_target_paralog(toronto_multi_target_paralog$multi_target_paralog_guides, 
                                                        toronto_multi_target_paralog$multi_target_not_paralog_guides, 
                                                        toronto_not_paralog$normal_sgrna_paralog, 
                                                        toronto_not_paralog$normal_sgrna_not_paralog,
                                                        toronto)

# Yusa paralog analysis
yusa_multi_target_paralog <- find_multi_target_paralog(yusa_multi_target_annotation$aln_data_df_1, predicted_paralog_pairs, 3)

yusa_not_paralog <- find_not_multi_target_paralog(yusa, yusa_multi_target_annotation$aln_data_df_1, '../barbara_36K_paralog_pairs.csv')

dataset_yusa_fischer <- fisher_multi_target_paralog(yusa_multi_target_paralog$multi_target_paralog_guides, 
                                                        yusa_multi_target_paralog$multi_target_not_paralog_guides, 
                                                        yusa_not_paralog$normal_sgrna_paralog, 
                                                        yusa_not_paralog$normal_sgrna_not_paralog,
                                                        yusa)

# Avana paralog analysis
avana_multi_target_paralog <- find_multi_target_paralog(avana_multi_target_annotation$aln_data_df_1, predicted_paralog_pairs, 3)

avana_not_paralog <- find_not_multi_target_paralog(avana, avana_multi_target_annotation$aln_data_df_1, '../barbara_36K_paralog_pairs.csv')

dataset_avana_fischer <- fisher_multi_target_paralog(avana_multi_target_paralog$multi_target_paralog_guides, 
                                                        avana_multi_target_paralog$multi_target_not_paralog_guides, 
                                                        avana_not_paralog$normal_sgrna_paralog, 
                                                        avana_not_paralog$normal_sgrna_not_paralog,
                                                        avana)


```

### Are genes with paralogs more likely to be subject to multi-targeting effects?

```{r}

fisher_multi_target_paralog_genes <- function(off_target_paralog_genes, off_target_not_paralog_genes, normal_paralog_genes, normal_not_paralog_genes, sgrna_library) {

  print(paste('Paralog genes targeted by multi-target sgRNAs:', length(off_target_paralog_genes)))
  print(paste('Non paralog genes targeted by multi-target sgRNAs:', length(off_target_not_paralog_genes)))
  print(paste('Paralog genes not targeted by multi-target sgRNAs:', length(normal_paralog_genes)))
  print(paste('Non paralog genes not targeted by multi-target sgRNAs:', length(normal_not_paralog_genes)))

  # Create the contingency table
  contingency_table <- matrix(
    c(
      length(off_target_paralog_genes),  # Multi-target sgRNAs targeting paralog genes
      length(off_target_not_paralog_genes),  # Multi-target sgRNAs not targeting paralog genes
      length(normal_paralog_genes),  # Non-multi-target sgRNAs targeting paralog genes
      length(normal_not_paralog_genes)  # Non-multi-target sgRNAs not targeting paralog genes
    ),
    nrow = 2
  )

  print(contingency_table)

  # Check if the total matches the unique sgRNAs in alignment_data
  total_contingency_table <- sum(contingency_table)
  total_unique_gene <- length(unique(sgrna_library$gene))

  print(paste('total sgRNA count in contingency table:', total_contingency_table))
  print(paste('total original number of gene:', total_unique_gene))

  if (total_contingency_table == total_unique_gene) {
    print("The sum of the contingency table matches the total unique gene in sgRNA library.")
  } else {
    print("Mismatch detected between the contingency table sum and the total unique gene in sgRNA library")
  }

  # Perform Fisher's test
  fischer_library_multi_target <- fisher.test(contingency_table)
  print(fischer_library_multi_target)

}

brunello_gene_paralog_multi_target <- unique(brunello_multi_target_paralog$aln_multi_target_paralog_df$gene)

brunello_gene_not_paralog_multi_target <- setdiff(
  unique(sub("^sg(.*?)_.*", "\\1", brunello_multi_target_paralog$multi_target_not_paralog_guides)),
  brunello_gene_paralog_multi_target
)

gene_paralog_target <- function(sgrna_library, gene_paralog_multi_target, gene_not_paralog_multi_target, barbara_paralog_dataset) {
  
  # Load paralog gene pairs
  barbara_paralog <- read_csv(barbara_paralog_dataset)
  
  # Extract the vector of paralog genes from column A2
  paralog_genes <- pull(barbara_paralog, A2)
  
  # Genes that are paralogs but not multi-target or not in the not-paralog list
  gene_paralog_normal_target <- sgrna_library %>%
    filter(
      gene %in% paralog_genes,
      !gene %in% gene_paralog_multi_target,
      !gene %in% gene_not_paralog_multi_target
    )
  
  # Genes that are NOT paralogs, and also not multi-target or in not-paralog list
  gene_not_paralog_normal_target <- sgrna_library %>%
    filter(
      !gene %in% paralog_genes,
      !gene %in% gene_paralog_multi_target,
      !gene %in% gene_not_paralog_multi_target
    )
  
  return(list(
    paralog_normal_target = unique(gene_paralog_normal_target$gene),
    not_paralog_normal_target = unique(gene_not_paralog_normal_target$gene)
  ))
}

brunello_paralog_normal <- gene_paralog_target(brunello, brunello_gene_paralog_multi_target, brunello_gene_not_paralog_multi_target, '../barbara_36K_paralog_pairs.csv')

brunello_gene_fisher <- fisher_multi_target_paralog_genes(brunello_gene_paralog_multi_target, 
                                                          brunello_gene_not_paralog_multi_target,
                                                          brunello_paralog_normal$paralog_normal_target,
                                                          brunello_paralog_normal$not_paralog_normal_target,
                                                          brunello)


```

```{r}
toronto_gene_paralog_multi_target <- unique(toronto_multi_target_paralog$aln_multi_target_paralog_df$gene)

toronto_gene_not_paralog_multi_target <- setdiff(
  unique(sub("^sg(.*?)_.*", "\\1", toronto_multi_target_paralog$multi_target_not_paralog_guides)),
  toronto_gene_paralog_multi_target
)

toronto_paralog_normal <- gene_paralog_target(toronto, toronto_gene_paralog_multi_target, toronto_gene_not_paralog_multi_target, '../barbara_36K_paralog_pairs.csv')

toronto_gene_fisher <- fisher_multi_target_paralog_genes(toronto_gene_paralog_multi_target, 
                                                          toronto_gene_not_paralog_multi_target,
                                                          toronto_paralog_normal$paralog_normal_target,
                                                          toronto_paralog_normal$not_paralog_normal_target,
                                                          toronto)


```



